---
output: github_document
---

<!-- README.md is generated from README.Rmd. Please edit that file -->

```{r, include = FALSE}
knitr::opts_chunk$set(
  collapse = TRUE,
  comment = "#>",
  fig.path = "man/figures/README-",
  out.width = "100%"
)
devtools::load_all()
```

# ordr.extra

<!-- badges: start -->
[![Lifecycle: experimental](https://img.shields.io/badge/lifecycle-experimental-orange.svg)](https://lifecycle.r-lib.org/articles/stages.html#experimental)
<!-- badges: end -->

This is an extension of [the **ordr** package](https://cran.r-project.org/package=ordr), a **tidyverse** extension for managing ordination models and rendering biplots.
**ordr** provides methods for handling only some of the most common techniques, so **ordr.extra** provides methods for several additional techniques.

This package is like **broom** in that it can expand to accommodate additional models until they exceed the bandwidth of its maintainer(s).
See [this issue](https://github.com/corybrunson/ordr.extra/issues/2) for additional discussion, and please comment with any additional suggestions!

## installation

**ordr.extra** is not yet on CRAN.
You can install the development version as follows:

``` r
# install.packages("devtools")
devtools::install_github("corybrunson/ordr.extra")
```

Note:
This package extends tools from **ordr** to classes of ordination models from several additional packages, but _it does not depend on these packages_.
The reason is so that someone interested in working with, say, **candisc** models does not have to also install **ca**, **PMA**, and the other packages **ordr.extra** supports.
The trade-off is that this user must manually install **candisc** as well.

## example

Joint correspondence analysis[^jca], implemented in the **ca** package, essentially applies PCA to a high-dimensional contingency table reformatted as a data frame with one observation per row.
To illustrate the technique, recall the UC Berkeley admissions data set, reformatted here as a data frame of counts:

[^jca]: Greenacre MJ (2005) "From Correspondence Analysis to Multiple and Joint Correspondence Analysis". Available at SSRN: <https://ssrn.com/abstract=847664>

```{r data}
# inspect admissions data
head(as.data.frame(UCBAdmissions))
```

We can use **ordr** syntax to model these data using joint correspondence analysis with the function `ca::mjca()`[^array], once we've ensured that **ca** is installed:

[^array]: Note that the array is being passed as input. `ca::mjca()` can also take data frame input, but it requires a format that differs from that produced by `as.data.frame()`.

```{r jca, warning=FALSE}
# install {ca} if not already installed
if (! "ca" %in% rownames(installed.packages())) install.packages("ca")
# fit JCA model to admissions data
(admissions_jca <- ordinate(UCBAdmissions, ca::mjca, lambda = "JCA"))
```

We can then generate a monoplot[^monoplot] of the group masses for each of the three categorical variables:

[^monoplot]: This is a monoplot because JCA is based on the SVD of a case-by-variable matrix, and only variable elements are plotted here.

```{r biplot, fig.height=3}
# build biplot of admissions JCA model with variance distributed to variables
admissions_jca %>%
  confer_inertia("colprincipal") %>%
  ggbiplot() +
  theme_bw() + theme_biplot() +
  geom_origin() +
  geom_cols_point(aes(color = factor, shape = factor, size = mass)) +
  geom_cols_text_repel(aes(label = level, color = factor), show.legend = FALSE) +
  scale_color_brewer(palette = "Dark2") +
  scale_size_area(guide = "none") +
  labs(color = "Factor level", shape = "Factor level")
```

## acknowledgments

See [the **ordr** repo](https://github.com/corybrunson/ordr) for full acknowledgments.
