---
title: "Extending ordr: PCA"
output: 
  rmarkdown::html_vignette:
    toc: TRUE
vignette: >
  %\VignetteIndexEntry{Extending ordr: PCA}
  %\VignetteEngine{knitr::rmarkdown}
  %\VignetteEncoding{UTF-8}
---

```{r, setup, include=FALSE}
knitr::opts_chunk$set(
  comment = '', fig.width = 6, fig.height = 6
)
```

# Introduction

This is the first of 3 vignettes that detail the process of extending **ordr** to incorporate new ordination methods. This vignette focuses on principal component analysis, a straightforward SVD-based technique, while the sequel focuses on factor analysis, a much more complicated case. We demonstrate how to extract key elements from each class based on a shared underlying mathematical theory and to make these available to **ordr** using a set of methods based on this theory. We also illustrate how unit tests and examples are written to complete the contribution. The final vignette applies both contributions to a common data set to showcase their commonalities and differences. The vignettes are intended for R users with experience in geometric data analysis and with the tidyverse.

If an extension of **ordr** to a new ordination class is agreed upon, the first step will be to open an issue that includes the checklist from this developer vignette: `vignette("extension-checklist")`.

```{r, message=FALSE}
library(ordr)
library(psych)
```

# Principal Component Analysis

Principal component analysis (PCA) is a geometric statistical method by which multivariate data is reduced to fewer dimensions, expending as little variance as possible in the process---that is, the resulting dimensions, called _principal axes_, contain as much of the variance as possible. The goal is to more easily visualize and interpret patterns in the data. PCA is useful in various fields; this vignette will conclude with an example using psychometric data.

## EVD and SVD

PCA can be done by applying either of two matrix decomposition methods to a centered and (often) scaled data matrix $\bar{X}_{n\times p}$.

One method, used by `stats::prcomp()`, is singular value decomposition (SVD).
Assuming $n > p$, SVD obtains the decomposition
$$\bar{X}_{n\times{p}} = U_{n\times{p}}D_{p\times{p}}(V_{p\times{p}})^\top\text{,}$$
where the column vectors of $U$ and of $V$ have length $1$ and are pairwise perpendicular and $D$ is diagonal.
The columns of $U$ are called the _left singular vectors_ of $\bar{X}$ and those of $V$ the _right singular vectors_; the diagonal elements of $D$ are the _singular values_.
Based on the equation, elements from $U$ and from $V$, respectively, are referred to as "row" and "column" elements.

```{r}
scaled_iris <- scale(iris[1:4])
pca_stats <- prcomp(scaled_iris, rank. = 4)
```

The other method uses eigenvalue decomposition (EVD), or just eigendecomposition, as in `stats::princomp()` and in `psych::principal()`. The EVD of a full-rank square matrix $M$ is the unique decomposition
$$M = V \Lambda V^\top$$
with $\Lambda$ diagonal; the entries of $\Lambda$ are called the eigenvalues, and the columns of $V$ are called the eigenvectors, of $M$.
For PCA, we obtain $U$ as the matrix of eigenvectors of $\bar{X}\bar{X}^\top$ and $V$ as the that of $\bar{X}^\top\bar{X}$. The positive eigenvalues from these two decompositions agree, and their square roots are the singular values of $\bar{X}$. The fold demonstrates these connections between SVD and EVD.

```{r}
( pca_psych <- principal(scaled_iris, nfactors = 4, rotate = "none") )
```

<details>
<summary> Derive EVD from SVD </summary>

To understand EVD in terms of SVD, again take $X=UDV^\top$. Then when we consider the covariance matrix $\bar{X}^\top \bar{X}$, we have

$$
\begin{align*}
(\bar{X}_{n\times{p}})^\top \bar{X}_{n\times{p}} &= (U_{n\times{p}}D_{p\times{p}}(V_{p\times{p}})^\top )^\top  U_{n\times{p}}D_{p\times{p}}(V_{p\times{p}})^\top \tag{1}\\
&= ((V_{p\times{p}})^\top )^\top D_{p\times{p}}^\top (U_{n\times{p}})^\top U_{n\times{p}}D_{p\times{p}}(V_{p\times{p}})^\top \tag{2}\\
&= V_{p\times{p}}D_{p\times{p}}D_{p\times{p}}(V_{p\times{p}})^\top \tag{3}\\
&= V_{p\times{p}}(D_{p\times{p}})^2(V_{p\times{p}})^\top \tag{4}.
\end{align*}
$$

Note: we use the orthogonality of $U$ to obtain (3) as $U^\top = U^{-1}$.

</details>

In reference to factor analysis, $V$ is often called the loadings matrix---the columns contain the linear coefficients of the measured variables as they load onto each principal component---while $U$ contains the scores.
<!--In EVD, $\bar{X}$ must be multiplied into the decomposition to obtain scores, demonstrated later.-->
Depending on whether the investigator's focus is on the variables or the cases, they may want to study and visualize the loadings or the scores, respectively. When focusing on the scores, the row data (the cases) may be studied using _principal coordinates_, the rows of $UD$, while the column data (the variables) are left in _standard coordinates_, the rows of $V$. Here, the inertia of $D$ has been _conferred_ onto the left singular vectors, changing the space in which they are being coordinatized.
When focusing instead on the variables, the inertia will often be conferred onto the right singular vectors, putting the variables in principal coordinates (the rows of $VD$) and the cases in standard coordinates (the rows of $U$).

## Necessary Recovery Methods

**TODO: I'm not sure these need to be introduced here; they could instead be introduced one or two at a time in the coming sections. --JCB**

By *recovery methods* or *recoverers* we mean the functions that retrieve key elements from an ordination object. The essential recoverers extract the components of the decomposition just discussed, while others extract supplementary elements and annotations specific to the technique and its implementation:

1. `ordr::recover_inertia()` returns the eigenvalues of our EVD, or equivalently the squared singular values of our SVD---the diagonal entries of $D^2$.
2. `ordr::recover_coord()` returns the names of the coordinate axes of our ordination object.
3. `ordr::recover_conference()` indicates how the inertia is distributed to the rows or columns ($U$ or $V$) in the object. (**stats** and **psych** make different choices here.)
4. `ordr::recover_rows()` returns $U$, or more generally $UD^a$ for some $a$, which factors into the SVD from the left.
5. `ordr::recover_cols()` returns $V$ or some $VD^b$, whith factors in from the right.
6. `ordr::recover_supp_rows()` returns any row elements associated with the left factor of the SVD, in a sense to be defined later.
7. `ordr::recover_supp_cols()` returns any column elements associated with the right factor.
8. `ordr::recover_aug_rows()` and (9) `ordr::recover_aug_cols()` do not recover raw elements of the PCA but rather reassemble additional, often non-numeric elements that augment the row and column coordinates.

**This makes sense to introduce here, as a goal toward which we're working. --JCB**

While these recovery methods will depend on the GDA technique, the helper function `ordr::as_tbl_ord()` reduces redundancy and expedites debugging, so we give it a method for the `principal` class:

```{r}
as_tbl_ord.principal <- getFromNamespace("as_tbl_ord_default", "ordr")
```

This method draws from the recoverers to produce a `tbl_ord`, illustrated below with the pre-loaded set of Anderson's iris data:

```{r}
( tbl_ord <- as_tbl_ord(pca_stats) )
```

## Row and Column Elements

In an SVD $\bar{X} = UDV^\top$, the left and right matrix factors contain coefficients of the row elements and column elements, respectively, of the original data $X$ in a new coordinate space. This is why we refer to elements of the left factor $U$ (or $UD$) as "row elements" and those of the right factor $V$ (or $VD$) as "column elements". With SVD, both factors are said to contain "active" elements.

An EVD-based method might be based on the row inner products $X X^\top$ or on the column inner products $\bar{X}^\top \bar{X}$. Either way, the EVD yields a symmetric factorization $V \Lambda V^\top = (VD)(VD)^\top$; the "left" and "right" factors are the same.
EVD-based PCA uses the latter, a decomposition of the covariance matrix $X^\top X$ (which happily means that the $V$ in this paragraph is the same as that in the last). We therefore interpret the loadings $VD$ as active column elements, and we write the column recovery method to extract this matrix:

```{r}
recover_cols.principal <- function(x) {
  unclass(x[["loadings"]])
}
recover_cols.principal(pca_psych)
```

We obtain no active row elements from this class, so we write the row recovery method to return a matrix with no rows (elements) but the correct number and names of the columns (artificial coordinates):

```{r}
recover_rows.principal <- function(x) {
  matrix(
    nrow = 0, ncol = ncol(x[["loadings"]]),
    dimnames = list(NULL, colnames(x[["loadings"]]))
  )
}
recover_rows.principal(pca_psych)
```

We do obtain case scores, which are unambiguously row elements; however, they are not obtained through SVD but computed internally. We can still verify that they multiply with the loadings to reconstruct $\bar{X}$:

<details>
<summary> Check that scores and loadings reconstruct data </summary>

We divide the scores--loadings product elementwise by the data matrix and print the first few rows:

```{r}
head(
  (pca_psych$scores %*% t(pca_psych$loadings)) / scaled_iris
)
```

We use element-wise division to show equality rather than `all.equal()` or `identical()` because rounding errors may result in small differences in computation despite mathematical equivalence.
The matrix of all 1s indicates that the original matrices were equal.

</details>

Hence, the scores are supplementary row elements, and we can go ahead and write a recovery method for them. We must also account for cases where `scores = FALSE` is passed, resulting in no `scores` element in the `principal`-class output.

```{r}
recover_supp_rows.principal <- function(x) {
  if (is.null(x[["scores"]])) {
    tibble(numeric(0), nrow = 0, ncol = ncol(x[["loadings"]]))
  }
  else {
    x[["scores"]]
  }
}
head(recover_supp_rows.principal(pca_psych))
```

## Inertia and its Conference

From the documentation, `psych::principal()` returns the eigenvalues in the `values` element of the output:

```{r}
pca_psych$values
```

We can validate these values against those returned using the recovery method for the `prcomp` class:

```{r}
recover_inertia(pca_stats) / pca_psych$values
```

Notice that $149=n-1$, these eigenvalues have been scaled by the degrees of freedom. To obtain the inertia, we need then to multiply them by $n-1$:

```{r}
recover_inertia.principal <- function(x) {
  x[["values"]] * (nrow(x[["scores"]]) - 1L)
}
recover_inertia.principal(pca_psych)
```

This method is consistent with that for the `prcomp` class.

These values are usually given the names of the artificial coordinates. Because these names are used for several other purposes, they have a separate recovery method. As with the row element recoverers above, we obtain these names from the column names of the loading matrix:

```{r}
recover_coord.principal <- function(x) {
  colnames(x[["loadings"]])
}
recover_coord.principal(pca_psych)
```

The inertia is established, but the methods must also account for how it is conferred on the matrix factors. This "conference" governs the geometry of biplots and matters crucially to their interpretation, but the matrix factors returned by different techniques may confer it differently. To avoid having to manipulate the matrix factors during recovery, we use the `recover_conference()` method to inform **ordr** how this conference is done.

<details>
<summary> Implications of different conferences of inertia </summary>

**Should this be trimmed or cut given the additional text above, or should the additional text be folded in here? (That is to ask whether the un-folded text about the meaning and importance of conference of inertia is too technical.) --JCB**

Not distributing the inertia (i.e. not multiplying out the matrix $D$ in SVD or EVD) maintains $U$ and $V$ as matrices of unit vectors. Thus, we say that the cases and variables are in standard coordinates. But when we _do_ distribute the inertia to one or both other matrices in the decomposition, we obtain the cases and/or variables in principal coordinates.

Assuming we desire principal coordinates, EVD-based techniques leave only the option of conferring inertia onto one matrix of singular vectors. This depends on the implementation (and how we interpret it), and in the case of `psych::principal()` it would be the right singular vectors. However, the conference of inertia in SVD-based techniques depends on the user's motivation. In short, if the user wishes to maintain the approximate Euclidean distances between data cases, then conference onto the left singular vectors is ideal. If instead the user wishes to maintain the approximate correlations between variables, then conference onto the right singular vectors is ideal.

```{r}
recover_conference(pca_stats)
```

The conference recovery method is, so far, always a static vector, since the methods don't confer inertia differently based on the input. Here, we see that `stats::prcomp()` automatically confers all inertia onto the left singular vectors (a conference of $(0,1)$, alternatively, would indicate that a PCA has conferred its inertia onto the right singular vectors). This means that the scores of `pca_stats` are in principal coordinates and the loadings are in standard coordinates.

</details>

This is straightforward for the loadings: We expect from the EVD that they are in principal coordinates, i.e. that they have full inertia, and we checked that the loadings matrix agrees with the full-inertia loadings obtained from `stats::prcomp()`. So, $L = VD$. We also checked that the product $SL^\top$ reconstructs $\bar{X}$, which suggests that $S = U$. So, the scores should be in standard coordinates, i.e. they have been conferred no inertia. Symbolically:

$$ X = UDV^\top = (UD^0)(VD^1)^\top = SL^\top$$

This yields a conference of $(0,1)$. A careful check follows.

<details>
<summary> Check that the loadings matrix has full inertia </summary>

```{r}
# X'X from EVD
cov <- cov(scaled_iris)
evd <- eigen(cov)
# observe loadings matrix equals VD up to sign
( evd$vectors %*% diag(sqrt(evd$values))) / unclass(pca_psych$loadings )
```

Hence we have $L_{p\times{p}} = V_{p\times{p}}D_{p\times{p}}$.

</details>

As such, we define the following recovery method:

```{r}
recover_conference.principal <- function(x) {
  c(0, 1)
}
recover_conference.principal(pca_psych)
```

To illustrate the entire decomposition, we have the following code.

<details>
<summary>Show matrix algebra validation</summary>

**Was this supposed to be a different check? It seems to check only that the `evd()` function does what it should. --JCB**

```{r}
cov / (evd$vectors %*% diag(evd$values) %*% t(evd$vectors))
```

As indicated by the matrix of all $1$s, the covariance matrix and product of matrices are equal.

</details>

## PCA Followed by Rotation

In factor analysis, the loadings---a transformation of the measured or "manifest" variables into a smaller set of latent variables---will often be composed with a rotation---a change of basis to a new but same-sized set of latent variables---to achieve "simple structure" in which the latent variables are easier to interpret. We'll discuss this in more detail in the FA vignette. Because PCA has long been appropriated for FA, such rotations may also be applied to the results of PCA, and this functionality is provided by **psych**.

One way **ordr** might organize this information would be to separately recover the unrotated loadings and the rotation matrix. This would require some transformation, since the `principal` class stores the already-rotated loadings, but it might be worth the trouble to disentangle other aspects of the class and to generalize the recovery process to more methods. For now, however, **ordr** recovers elements for the rotated model, and this requires additional caution and checks when writing the methods.

Rotations may be either orthogonal or oblique. We compute and illustrate both in the next two code chunks, using an oblique rotation provided by **GPArotation**.

<details>
<summary> Remark about precision of language </summary>

Mathematically, oblique rotations are not proper rotations, which are angle-preserving. Following the FA literature, we refer to both as "rotations" but always qualify those that are oblique.

Fundamental to PCA are the orthogonality of the principal components and the minimality of the residual variance of each principal subspace. Oblique rotations compromise the former, while even orthogonal rotations undermine the latter. For these reasons, it is contested whether PCA with rotation should still be considered PCA. For clarity, we refer to these models as "PCA _followed by_ rotation".

</details>

```{r, message=FALSE}
library(GPArotation)
( pca_ortho <- principal(scaled_iris, rotate = "varimax", nfactors = 4) )
( pca_obliq <- principal(scaled_iris, rotate = "oblimin", nfactors = 4) )
```

```{r, warning=FALSE, fig.width=8, fig.height=3}
op <- par(mfrow = c(1, 3))
biplot.psych(pca_psych, choose = c(1, 2), main = "no rotation")
biplot.psych(pca_ortho, choose = c("RC1","RC2"), main = "orthogonal rotation")
biplot.psych(pca_obliq, choose = c("TC1","TC2"), main = "oblique rotation")
par(op)
```

The biplots suggest that the rotations twist the manifest variables (vectors) to better align with the latent variables (axes); in fact, they twist the latent variables to better align with the manifest variables. Because all 4 factors (principal components) were obtained, the rotations are not constrained to two dimensions, and the geometry of the vectors in the 2-dimensional projection warps. However, both rotations preserve the correlations between the manifest variables in the transformed space. (The angles between vectors appear to change dramatically under the oblique rotation, but recall that the rotated principal axes are no longer orthogonal; biplot does not communicate this.)

**RESUME.**

Now let us compare the categories of rotation by examining the linear algebra at play. In PCA followed by an orthogonal rotation, our loadings are the original loadings matrix $L_{p\times{p}}$ multiplied by the orthogonal rotation matrix $T_{p\times{p}}$.

<details>
<summary>Show matrix equality validation</summary>
```{r}
pca_ortho$loadings[,c("RC1", "RC2", "RC3", "RC4")] / (pca_psych$loadings %*% pca_ortho$rot.mat) # matrices are equal up to sign
```
</details>

But in the oblique case, we have two loadings matrices. The first is the *pattern matrix* $L_{p\times{p}}$, which represents the loadings as we first defined them (i.e. the coefficients in our linear combination of variables that forms each principal component, prior to any rotation). The second is the *structure matrix* $S_{p\times{p}} = L_{p\times{p}}T_{p\times{p}}$. Below, we demonstrate that the extractable loadings element from an oblique-transformed `principal` object is the structure matrix.

<details>
<summary>Show matrix equality validation</summary>
```{r}
pca_obliq$loadings[,c("TC1", "TC2", "TC3", "TC4")] / (pca_psych$loadings %*% pca_obliq$rot.mat) # LT equals the extracted loadings matrix up to sign
```
</details>

<details>
<summary> Interpretation of pattern versus structure matrices </summary>
The pattern matrix is useful when we want to consider the principal components apart from their correlations with one another. It directly explains how much each principal component loads into each variable. On the other hand, the structure matrix contains correlations between factors and variables, and it tends to be more stable between samples. Notably, the structure matrix can be obtained from the pattern matrix by multiplying the latter by $\Phi_{p\times p}$, the interfactor correlation matrix.

<details>
<summary>Show matrix equality validation</summary>
```{r}
(pca_obliq$loadings %*% pca_obliq$Phi) / pca_obliq$Structure # matrix of all 1s indicates equality
```
</details>

This is why we have just one loadings matrix in the orthogonal case: with an orthogonal rotation, $\Phi_{p\times p}$ is the identity matrix.
</details>

<details>
<summary> How to decide between orthogonal and oblique rotation </summary>
When choosing between an orthogonal and oblique rotation following PCA, the right choice depends on what we wish to prioritize. An orthogonal rotation keeps more faithful to the data. But if we want to convey that there is significant correlation in the variables, or if we want a more interpretable biplot, then an oblique rotation may be more appropriate. Moreover, various different rotations suitable for different goals exist within either category.
</details>

For **ordr**, we use the same recovery method for rotated PCAs as for unrotated. Furthermore, we opt against adding recovery methods for the rotation matrices, the motivation being that rotations do not necessarily uphold key characteristics of PCA (namely orthogonality and maximal captured variance). Because of this, it is debated whether PCA followed by rotation/transformation is PCA at all; thus, it may be misguided to take the rotation matrix as an element of PCA for **ordr**.

```{r}
recover_cols.principal(pca_obliq)
recover_cols.principal(pca_ortho)
```

## Augmentation

At this point, we have recovered almost all of the elements from our PCA. The last elements to recover will be annotations to the matrix factors, including the inertia.

Now, the last step in writing `psych::principal()` methods into **ordr** is the augmentation functions that are essential to inspecting and understanding a `tbl_ord` object. Essentially, the purpose of these functions is to place a name to the various PCA elements that have been extracted and join these elements with information that is not part of the matrix decomposition. The augmented matrices are where the user should see whether PCA elements come from the rows or columns and whether they are active or supplementary. Moreover, they should offer the user measures such as the mean and standard deviation when available from the original function output. We write the following functions accordingly.

Starting with `recover_aug_rows.principal()`, our function should be defined to indicate that no row elements are active, and that the supplementary elements are scores *if* a scores object exists.

```{r}
recover_aug_rows.principal <- function(x) {
  res <- tibble(.rows = 0L)
  
  # scores as supplementary points
  if (!is.null(x[["scores"]])) {
    name <- rownames(x[["scores"]])
    res_sup <- if (is.null(name)) {
      tibble(.rows = nrow(x[["scores"]]))
    } else {
      tibble(name = name)
    }
  } else {
    res_sup <- tibble(.rows = 0L)
  }
  
  # supplement flag
  res$.element <- "active"
  res_sup$.element <- "score"
  as_tibble(dplyr::bind_rows(res, res_sup))
}
```

We must also be sure that the number of rows in the augmented matrix agree with those of the extracted matrix.

<details>
<summary> Verification of number of rows </summary>
```{r}
library(tibble) # load {tibble} in order to use `ordr::as_tbl_ord()` below

as_tbl_ord(pca_psych) |>
  recover_aug_rows.principal() |>
  nrow()
```
</details>

The scores constitute a $150 \times 4$ matrix, consistent with the dimensions here. We similarly define and validate `recover_aug_cols.principal()`. Here, we will augment the column vectors with attributes `center` and `scale` since these values correspond to the variables and are required for some biplot layers like calibrated axes.

```{r}
recover_aug_cols.principal <- function(x) {
  name <- rownames(x[["loadings"]])
  res <- if (is.null(name)) {
    tibble(.rows = nrow(x[["loadings"]]))
  } else {
    tibble(name = name)
  }
  res$.element <- "active"
  res
}
```

<details>
<summary> Verification of number of rows </summary>
```{r}
ordr:::as_tbl_ord(pca_psych) |>
  recover_aug_cols.principal() |>
  nrow()
```
</details>

Last, we write `recover_aug_coord.principal()` to add coordinate names and a column of standard deviations (which are directly provided in the output of `psych::principal()`, so users may expect them) to the loadings element we recovered. Again, we will verify that the number of rows is what we would expect.

```{r}
recover_aug_coord.principal <- function(x) {
  data.frame(
    name = recover_coord.principal(x),
    sdev = sqrt(x[["values"]][seq(1, ncol(x[["loadings"]]))])
  )
}
```

<details>
<summary> Verification of number of rows </summary>
```{r}
as_tbl_ord(pca_psych) |>
  recover_aug_coord.principal() |>
  nrow()
```
</details>

With these functions defined, we can obtain the `tbl_ord` object below in two concise ways, either by making our existing "principal" object a `tbl_ord` or by generating the `tbl_ord` from scratch.

```{r}
as_tbl_ord(pca_psych) |>
  augment_ord()

ordinate(scaled_iris, ~ principal(., nfactors = 4, rotate = "none"))
```

## Examples

With all of our recoverers written and validated, we finalize the contribution to **ordr.extra** by defining and documenting the recoverers in a "methods" file following the same pattern as `methods-stats-prcomp.r`.

Then, we need to write an "examples" script following the same pattern as `ex-methods-prcomp-iris.r`. The example looks as follows:

```{r}
# data frame of Anderson iris species measurements
class(iris)
head(iris)

if (require(psych)) {# {psych}

# compute unscaled row-principal components of scaled measurements
iris[, -5] |>
  psych::principal(nfactors = 4, rotate = "none") |>
  as_tbl_ord() |>
  print() -> iris_pca

# recover observation principal coordinates and measurement standard coordinates
head(get_rows(iris_pca))
get_cols(iris_pca)

# augment measurement coordinates with names and scaling parameters
(iris_pca <- augment_ord(iris_pca))

}# {psych}
```

## Methods

The last file to complete our contribution is a unit test script. Again, we can follow the format of the analogous `test-stats-prcomp.r` and write the following:

```{r}
library(testthat) # load {testthat} to be used for each unit test

fit_principal <- psych::principal(iris[, -5], nfactors = 4, rotate = "none")

test_that("'principal' accessors have consistent dimensions", {
  expect_equal(ncol(get_rows(fit_principal)), ncol(get_cols(fit_principal)))
  expect_equal(ncol(get_rows(fit_principal)),
               length(recover_inertia(fit_principal)))
})

test_that("'principal' has specified distribution of inertia", {
  expect_type(recover_conference(fit_principal), "double")
  expect_vector(recover_conference(fit_principal), size = 2L)
})

test_that("'principal' augmentations are consistent with '.element' column", {
  expect_equal(".element" %in% names(recover_aug_rows(fit_principal)),
               ".element" %in% names(recover_aug_cols(fit_principal)))
})

test_that("`as_tbl_ord()` coerces 'principal' objects", {
  expect_true(valid_tbl_ord(as_tbl_ord(fit_principal)))
})
```

## Submission

The function `devtools::check()` is useful for verifying that the contribution will not generate errors or warnings. Once the extension has been formatted and checked, the last step is to submit a pull request following the guidelines of `CONTRIBUTING.md`.
