---
title: "Extending ordr: PCA"
output: 
  rmarkdown::html_vignette:
    toc: TRUE
vignette: >
  %\VignetteIndexEntry{extending-ordr-pca}
  %\VignetteEngine{knitr::rmarkdown}
  %\VignetteEncoding{UTF-8}
---

```{r, setup, include=FALSE}
knitr::opts_chunk$set(
  comment = '', fig.width = 6, fig.height = 6
)
```

# Introduction

This is the first of 3 vignettes that detail the process of extending **ordr** to incorporate new ordination methods. This vignette focuses on principal component analysis, a straightforward SVD-based technique, while the sequel focuses on factor analysis, a much more complicated case. We demonstrate how to extract key elements from each class based on a shared underlying mathematical theory and to make these available to **ordr** using a set of methods based on this theory. We also illustrate how unit tests and examples are written to complete the contribution. The final vignette applies both contributions to a common data set to showcase their commonalities and differences. The vignettes are intended for R users with experience in geometric data analysis and with the tidyverse.

If an extension of **ordr** to a new ordination class is agreed upon, the first step will be to open an issue that includes the checklist from this developer vignette: `vignette("extension-checklist")`.

```{r, message = FALSE}
library(ordr)
```

# Principal Component Analysis

Principal component analysis (PCA) is a geometric statistical method by which multivariate data is reduced to fewer dimensions, expending as little variance as possible in the process---that is, the resulting dimensions, called _principal axes_, contain as much of the variance as possible. The goal is to more easily visualize and interpret patterns in the data. PCA is useful in various fields; this vignette will conclude with an example using psychometric data.

## EVD and SVD

PCA can be done by applying either of two matrix decomposition methods to a centered and (often) scaled data matrix $\bar{X}_{n\times p}$.

One method, used by `stats::prcomp()`, is singular value decomposition (SVD).
Assuming $n > p$, SVD obtains the decomposition
$$\bar{X}_{n\times{p}} = U_{n\times{p}}D_{p\times{p}}(V_{p\times{p}})^\top\text{,}$$
where the column vectors of $U$ and of $V$ have length $1$ and are pairwise perpendicular and $D$ is diagonal.
The columns of $U$ are called the _left singular vectors_ of $\bar{X}$ and those of $V$ the _right singular vectors_; the diagonal elements of $D$ are the _singular values_.
Based on the equation, elements from $U$ and from $V$, respectively, are referred to as "row" and "column" elements.

The other method uses eigenvalue decomposition (EVD), or just eigendecomposition, as in `stats::princomp()` and in `psych::principal()`. The EVD of a full-rank square matrix $M$ is the unique decomposition
$$M = V \Lambda V^\top$$
with $\Lambda$ diagonal; the entries of $\Lambda$ are called the eigenvalues, and the columns of $V$ are called the eigenvectors, of $M$.
For PCA, we obtain $U$ as the matrix of eigenvectors of $\bar{X}\bar{X}^\top$ and $V$ as the that of $\bar{X}^\top\bar{X}$. The positive eigenvalues from these two decompositions agree, and their square roots are the singular values of $\bar{X}$. The fold demonstrates these connections between SVD and EVD.

<details>
<summary> Derivation of EVD from SVD </summary>
To understand EVD in terms of SVD, again take $X=UDV^\top$. Then when we consider the covariance matrix $\bar{X}^\top \bar{X}$, we have

$$
\begin{align*}
(\bar{X}_{n\times{p}})^\top \bar{X}_{n\times{p}} &= (U_{n\times{p}}D_{p\times{p}}(V_{p\times{p}})^\top )^\top  U_{n\times{p}}D_{p\times{p}}(V_{p\times{p}})^\top \tag{1}\\
&= ((V_{p\times{p}})^\top )^\top D_{p\times{p}}^\top (U_{n\times{p}})^\top U_{n\times{p}}D_{p\times{p}}(V_{p\times{p}})^\top \tag{2}\\
&= V_{p\times{p}}D_{p\times{p}}D_{p\times{p}}(V_{p\times{p}})^\top \tag{3}\\
&= V_{p\times{p}}(D_{p\times{p}})^2(V_{p\times{p}})^\top \tag{4}.
\end{align*}
$$

Note: we use the orthogonality of $U$ to obtain (3) as $U^\top = U^{-1}$.
</details>

In reference to factor analysis, $V$ is often called the loadings matrix---the columns contain the linear coefficients of the measured variables as they load onto each principal component---while $U$ contains the scores.
<!--In EVD, $\bar{X}$ must be multiplied into the decomposition to obtain scores, demonstrated later.-->
Depending on whether the investigator's focus is on the variables or the cases, they may want to study and visualize the loadings or the scores, respectively. When focusing on the scores, the row data (the cases) may be studied using _principal coordinates_, the rows of $UD$, while the column data (the variables) are left in _standard coordinates_, the rows of $V$. Here, the inertia of $D$ has been _conferred_ onto the left singular vectors, changing the space in which they are being coordinatized.
When focusing instead on the variables, the inertia will often be conferred onto the right singular vectors, putting the variables in principal coordinates (the rows of $VD$) and the cases in standard coordinates (the rows of $U$).

## Necessary Recovery Methods

**TODO: I'm not sure these need to be introduced here; they could instead be introduced one or two at a time in the coming sections. --JCB**

By *recovery methods* or *recoverers* we mean the functions that retrieve key elements from an ordination object. The essential recoverers extract the components of the decomposition just discussed, while others extract supplementary elements and annotations specific to the technique and its implementation:

1. `ordr::recover_inertia()` returns the eigenvalues of our EVD, or equivalently the squared singular values of our SVD---the diagonal entries of $D^2$.
2. `ordr::recover_coord()` returns the names of the coordinate axes of our ordination object.
3. `ordr::recover_conference()` indicates how the inertia is distributed to the rows or columns ($U$ or $V$) in the object. (**stats** and **psych** make different choices here.)
4. `ordr::recover_rows()` returns $U$, or more generally $UD^a$ for some $a$, which factors into the SVD from the left.
5. `ordr::recover_cols()` returns $V$ or some $VD^b$, whith factors in from the right.
6. `ordr::recover_supp_rows()` returns any row elements associated with the left factor of the SVD, in a sense to be defined later.
7. `ordr::recover_supp_cols()` returns any column elements associated with the right factor.
8. `ordr::recover_aug_rows()` and (9) `ordr::recover_aug_cols()` do not recover raw elements of the PCA but rather reassemble additional, often non-numeric elements that augment the row and column coordinates.

While these recovery methods will depend on the GDA technique, the helper function `ordr::as_tbl_ord()` reduces redundancy and expedites debugging, so we give it a method for the `principal` class:

```{r}
as_tbl_ord.principal <- getFromNamespace("as_tbl_ord_default", "ordr")
```

This method draws from the recoverers to produce a `tbl_ord`, illustrated below with the pre-loaded set of Anderson's iris data:

```{r}
scaled_iris <- scale(iris[1:4])
pca_stats <- prcomp(scaled_iris, rank. = 4)
( tbl_ord <- as_tbl_ord(pca_stats) )
```

## Conference of Inertia

The conference of inertia determines how the variance of our data is distributed. This will take effect in the geometry of PCA biplots. And because this conference directly impacts the row and column elements that we will need to recover, we ought to recover conference of inertia early on.

To this end, let us first recover the inertia itself, using the recovery method for `stats::prcomp()` as a benchmark against the `values` element of an analogous `principal` object (which is simply the eigenvalues of the input covariance matrix).

```{r}
library(psych) # load {psych} for use of PCA function below

pca_psych <- principal(scaled_iris, nfactors = 4, rotate = "none") # conduct PCA on the same X
recover_inertia(pca_stats) / pca_psych$values
```

Notice that $149=n-1.$ So, to obtain the inertia from a "principal" object, we need only to multiply the `values` element by $n-1$. Thus we define `recover_inertia.principal()` as follows.

```{r}
recover_inertia.principal <- function(x) {
  x[["values"]] * (nrow(x[["scores"]]) - 1)
}

recover_inertia.principal(pca_psych)
```
Now, our inertia is consistent with that recovered from the `prcomp` object.

With the inertia established, next we consider how it is conferred. 

<details>
<summary> Implications of different conferences of inertia </summary>
Not distributing the inertia (i.e. not multiplying out the matrix $D$ in SVD or EVD) maintains $U$ and $V$ as matrices of unit vectors. Thus, we say that the cases and variables are in standard coordinates. But when we *do* distribute the inertia to one or both other matrices in the decomposition, we obtain the cases and/or variables in principal coordinates.

Assuming we do wish to have principal coordinates, EVD-based techniques leave only the option of conferring inertia onto one matrix of singular vectors. This depends on the implementation (and how we interpret it), and in the case of `psych::principal()` it would be the right singular vectors. However, the conference of inertia in SVD-based techniques depends on the user's motivation. In short, if the user wishes to maintain the approximate Euclidean distances between data cases, then conference onto the left singular vectors is ideal. If instead the user wishes to maintain the approximate correlations between variables, then conference onto the right singular vectors is ideal.
</details>

```{r}
recover_conference(pca_stats)
```

The conference recovery method is, so far, always a static vector, since the methods don't confer inertia differently based on the input. Here, we see that `stats::prcomp()` automatically confers all inertia onto the left singular vectors (a conference of $(0,1)$, alternatively, would indicate that a PCA has conferred its inertia onto the right singular vectors). This means that the scores of `pca_stats` are in principal coordinates and the loadings are in standard coordinates.

When we recover the conference of inertia in `psych::principal()`, we would expect a $(0,1)$ conference given that `psych::principal()` is built around EVD (i.e. there are no rows $U$ to confer onto, only the columns $V^\top$). We can check this expectation as follows:

<details>
<summary>Show matrix algebra validation</summary>
```{r}
cov <- cov(scaled_iris) # X*X from EVD
evd <- eigen(cov)
(evd$vectors %*% diag(sqrt(evd$values))) / unclass(pca_psych$loadings) # observe loadings are equal to VD up to sign
```

Note that we use element-wise division to show equality rather than `all.equal()` or `identical()` because rounding errors may result in small differences in computation despite mathematical equivalence.

Hence, we have $L_{p\times{p}} = V_{p\times{p}}D_{p\times{p}}$, where $L$ denotes the loadings matrix. Since $L$ will be the matrix we recover for the right factor, this implies a $(0,1)$ conference. 
</details>

As such, we define the following recovery method.

```{r}
recover_conference.principal <- function(x) {
  c(0, 1)
}

recover_conference.principal(pca_psych)
```

To illustrate the entire decomposition, we have the following code.

<details>
<summary>Show matrix algebra validation</summary>
```{r}
cov / (evd$vectors %*% diag(evd$values) %*% t(evd$vectors))
```
As indicated by the matrix of all $1$s, the covariance matrix and product of matrices are equal.
</details>

## Row and Column Elements

As mentioned in the context of SVD, we can classify elements of PCA by what they "contribute" to the decomposition. That is, we can refer to those elements whose rows [columns] are multiplied into the decomposition as "row elements" ["column elements"]. Then when we consider EVD-based PCA, recall that EVD acts on the covariance matrix $X^\top X$, the column-wise inner product. Thus the loadings, $VD$, are active column elements.

```{r}
recover_cols.principal <- function(x) {
  unclass(x[["loadings"]])
}

recover_cols.principal(pca_psych)
```

But since we are drawing upon a column-wise inner product, we do not have an active row element in our PCA. Instead, we treat the scores as _supplementary_ elements. In EVD-based methods like `psych::principal()`, we can still obtain $X$ by multiplying the rows of scores by transposed loadings with full inertia. 

<details>
<summary>Show matrix algebra validation</summary>
```{r}
head((pca_psych$scores %*% t(pca_psych$loadings)) / scaled_iris) # matrix of all 1s indicates equality
```

</details>

Hence, the scores constitute our supplementary rows. And that $\bar{X}_{n \times p}$ is obtained by multiplying the scores by the loadings with full inertia shows the scores are in standard coordinates (i.e. no inertia has been conferred onto them).

```{r}
recover_rows.principal <- function(x) {
  matrix(nrow = 0, ncol = ncol(x[["loadings"]]),
         dimnames = list(NULL, colnames(x[["loadings"]])))
}

recover_rows.principal(pca_psych)

recover_supp_rows.principal <- function(x) {
  if (is.null(x[["scores"]])) {
    tibble(numeric(0), nrow = 0, ncol = ncol(x[["loadings"]]))
  }
  else {
    x[["scores"]]
  }
}

head(recover_supp_rows.principal(pca_psych))
```

Notice that in defining `recover_supp_rows.principal()`, we must also account for cases in which the `scores` element is null to avoid errors. This is necessary because PCA is applicable to covariance matrices, so we may not have data to obtain scores from.

## PCA Followed by Rotation

Next, we discuss the topic of rotations in PCA. Rotation is an optional step in PCA wherein our loadings matrix is multiplied by a rotation matrix. The motivation for doing so is to achieve a "simple structure" (i.e. a structure in which the latent variables behind our principal components provide an interpretable, explainable solution). We seek to extend **ordr** to work with these PCAs followed by rotation because, as we will demonstrate, rotations are changes of basis. Thus, the analysis of such a model can be completely different than that of the corresponding unrotated model. So, it makes sense to recover elements from the rotated model in its own right rather than to simply recover the rotation separately, then extract all the same elements as we would from the unrotated PCA.

<details>
<summary> Remark about precision of language </summary>
Before getting back into the details of extending **ordr**, we make a disclaimer about precision of language in regards to rotations. Because rotations can compromise the orthogonality of our principal components and because that orthogonality is a fundamental aspect of PCA, it is contested that PCA with rotation is still PCA. Thus, we will refer to these solutions as "PCA *followed by* a rotation."
</details>

We classify rotations as either orthogonal or oblique. The prior necessarily preserves orthogonality between the principal axes; the latter does not. And (mathematically speaking), oblique rotations are not rotations at all since they are not angle-preserving.

```{r, warning=FALSE}
library(GPArotation) # load {GPArotation} to access oblique rotation below

pca <- principal(scaled_iris, rotate = "none", nfactors = 4)
pca_orth <- principal(scaled_iris, rotate = "varimax", nfactors = 4) # orthogonal example
pca_ob <- principal(scaled_iris, rotate = "oblimin", nfactors = 4) # oblique example

op <- par(mfrow = c(1, 3)) # place biplots side by side

biplot.psych(pca, choose = c(1,2))
biplot.psych(pca_orth, choose = c("RC1","RC2"))
biplot.psych(pca_ob, choose = c("TC1","TC2"))

par(op)
```

Here, we see analogous biplots for the same PCA followed by no rotation, followed by an orthogonal rotation, and followed by an oblique rotation, respectively. As illustrated by the biplots, both orthogonal and oblique rotations shift the variable vectors' endpoints among the rotated principal axes (since these endpoints are the rotated loadings). However, both categories of rotation preserve the correlation between variable vectors among the rotated principal axes (the angles between vectors appear to change in the oblique rotation biplot, but recall that the rotated PC1 and rotated PC2 are non-orthogonal, and the biplot does not communicate this).

Now let us compare the categories of rotation by examining the linear algebra at play. In PCA followed by an orthogonal rotation, our loadings are the original loadings matrix $L_{p\times{p}}$ multiplied by the orthogonal rotation matrix $T_{p\times{p}}$.

<details>
<summary>Show matrix equality validation</summary>
```{r}
pca_orth$loadings[,c("RC1", "RC2", "RC3", "RC4")] / (pca$loadings %*% pca_orth$rot.mat) # matrices are equal up to sign
```
</details>

But in the oblique case, we have two loadings matrices. The first is the *pattern matrix* $L_{p\times{p}}$, which represents the loadings as we first defined them (i.e. the coefficients in our linear combination of variables that forms each principal component, prior to any rotation). The second is the *structure matrix* $S_{p\times{p}} = L_{p\times{p}}T_{p\times{p}}$. Below, we demonstrate that the extractable loadings element from an oblique-transformed `principal` object is the structure matrix.

<details>
<summary>Show matrix equality validation</summary>
```{r}
pca_ob$loadings[,c("TC1", "TC2", "TC3", "TC4")] / (pca$loadings %*% pca_ob$rot.mat) # LT equals the extracted loadings matrix up to sign
```
</details>

<details>
<summary> Interpretation of pattern versus structure matrices </summary>
The pattern matrix is useful when we want to consider the principal components apart from their correlations with one another. It directly explains how much each principal component loads into each variable. On the other hand, the structure matrix contains correlations between factors and variables, and it tends to be more stable between samples. Notably, the structure matrix can be obtained from the pattern matrix by multiplying the latter by $\Phi_{p\times p}$, the interfactor correlation matrix.

<details>
<summary>Show matrix equality validation</summary>
```{r}
(pca_ob$loadings %*% pca_ob$Phi) / pca_ob$Structure # matrix of all 1s indicates equality
```
</details>

This is why we have just one loadings matrix in the orthogonal case: with an orthogonal rotation, $\Phi_{p\times p}$ is the identity matrix.
</details>

<details>
<summary> How to decide between orthogonal and oblique rotation </summary>
When choosing between an orthogonal and oblique rotation following PCA, the right choice depends on what we wish to prioritize. An orthogonal rotation keeps more faithful to the data. But if we want to convey that there is significant correlation in the variables, or if we want a more interpretable biplot, then an oblique rotation may be more appropriate. Moreover, various different rotations suitable for different goals exist within either category.
</details>

For **ordr**, we use the same recovery method for rotated PCAs as for unrotated. Furthermore, we opt against adding recovery methods for the rotation matrices, the motivation being that rotations do not necessarily uphold key characteristics of PCA (namely orthogonality and maximal captured variance). Because of this, it is debated whether PCA followed by rotation/transformation is PCA at all; thus, it may be misguided to take the rotation matrix as an element of PCA for **ordr**.

```{r}
recover_cols.principal(pca_ob)
recover_cols.principal(pca_orth)
```

## Augmentation

At this point, we have recovered almost all of the elements from our PCA. The last elements to recover will be annotations to the matrix factors, including the inertia. First, we recover the names of the coordinates.

```{r}
recover_coord.principal <- function(x) {
  colnames(x[["loadings"]])
}

recover_coord.principal(pca_psych)
```
Now, the last step in writing `psych::principal()` methods into **ordr** is the augmentation functions that are essential to inspecting and understanding a `tbl_ord` object. Essentially, the purpose of these functions is to place a name to the various PCA elements that have been extracted and join these elements with information that is not part of the matrix decomposition. The augmented matrices are where the user should see whether PCA elements come from the rows or columns and whether they are active or supplementary. Moreover, they should offer the user measures such as the mean and standard deviation when available from the original function output. We write the following functions accordingly.

Starting with `recover_aug_rows.principal()`, our function should be defined to indicate that no row elements are active, and that the supplementary elements are scores *if* a scores object exists.

```{r}
recover_aug_rows.principal <- function(x) {
  res <- tibble(.rows = 0L)
  
  # scores as supplementary points
  if (!is.null(x[["scores"]])) {
    name <- rownames(x[["scores"]])
    res_sup <- if (is.null(name)) {
      tibble(.rows = nrow(x[["scores"]]))
    } else {
      tibble(name = name)
    }
  } else {
    res_sup <- tibble(.rows = 0L)
  }
  
  # supplement flag
  res$.element <- "active"
  res_sup$.element <- "score"
  as_tibble(dplyr::bind_rows(res, res_sup))
}
```

We must also be sure that the number of rows in the augmented matrix agree with those of the extracted matrix.

<details>
<summary> Verification of number of rows </summary>
```{r}
library(tibble) # load {tibble} in order to use `ordr::as_tbl_ord()` below

as_tbl_ord(pca_psych) |>
  recover_aug_rows.principal() |>
  nrow()
```
</details>

The scores constitute a $150 \times 4$ matrix, consistent with the dimensions here. We similarly define and validate `recover_aug_cols.principal()`. Here, we will augment the column vectors with attributes `center` and `scale` since these values correspond to the variables and are required for some biplot layers like calibrated axes.

```{r}
recover_aug_cols.principal <- function(x) {
  name <- rownames(x[["loadings"]])
  res <- if (is.null(name)) {
    tibble(.rows = nrow(x[["loadings"]]))
  } else {
    tibble(name = name)
  }
  res$.element <- "active"
  res
}
```

<details>
<summary> Verification of number of rows </summary>
```{r}
ordr:::as_tbl_ord(pca_psych) |>
  recover_aug_cols.principal() |>
  nrow()
```
</details>

Last, we write `recover_aug_coord.principal()` to add coordinate names and a column of standard deviations (which are directly provided in the output of `psych::principal()`, so users may expect them) to the loadings element we recovered. Again, we will verify that the number of rows is what we would expect.

```{r}
recover_aug_coord.principal <- function(x) {
  data.frame(
    name = recover_coord.principal(x),
    sdev = sqrt(x[["values"]][seq(1, ncol(x[["loadings"]]))])
  )
}
```

<details>
<summary> Verification of number of rows </summary>
```{r}
as_tbl_ord(pca_psych) |>
  recover_aug_coord.principal() |>
  nrow()
```
</details>

With these functions defined, we can obtain the `tbl_ord` object below in two concise ways, either by making our existing "principal" object a `tbl_ord` or by generating the `tbl_ord` from scratch.

```{r}
as_tbl_ord(pca_psych) |>
  augment_ord()

ordinate(scaled_iris, ~ principal(., nfactors = 4, rotate = "none"))
```

## Examples

With all of our recoverers written and validated, we finalize the contribution to **ordr.extra** by defining and documenting the recoverers in a "methods" file following the same pattern as `methods-stats-prcomp.r`.

Then, we need to write an "examples" script following the same pattern as `ex-methods-prcomp-iris.r`. The example looks as follows:

```{r}
# data frame of Anderson iris species measurements
class(iris)
head(iris)

if (require(psych)) {# {psych}

# compute unscaled row-principal components of scaled measurements
iris[, -5] |>
  psych::principal(nfactors = 4, rotate = "none") |>
  as_tbl_ord() |>
  print() -> iris_pca

# recover observation principal coordinates and measurement standard coordinates
head(get_rows(iris_pca))
get_cols(iris_pca)

# augment measurement coordinates with names and scaling parameters
(iris_pca <- augment_ord(iris_pca))

}# {psych}
```

## Methods

The last file to complete our contribution is a unit test script. Again, we can follow the format of the analogous `test-stats-prcomp.r` and write the following:

```{r}
library(testthat) # load {testthat} to be used for each unit test

fit_principal <- psych::principal(iris[, -5], nfactors = 4, rotate = "none")

test_that("'principal' accessors have consistent dimensions", {
  expect_equal(ncol(get_rows(fit_principal)), ncol(get_cols(fit_principal)))
  expect_equal(ncol(get_rows(fit_principal)),
               length(recover_inertia(fit_principal)))
})

test_that("'principal' has specified distribution of inertia", {
  expect_type(recover_conference(fit_principal), "double")
  expect_vector(recover_conference(fit_principal), size = 2L)
})

test_that("'principal' augmentations are consistent with '.element' column", {
  expect_equal(".element" %in% names(recover_aug_rows(fit_principal)),
               ".element" %in% names(recover_aug_cols(fit_principal)))
})

test_that("`as_tbl_ord()` coerces 'principal' objects", {
  expect_true(valid_tbl_ord(as_tbl_ord(fit_principal)))
})
```

## Submission

The function `devtools::check()` is useful for verifying that the contribution will not generate errors or warnings. Once the extension has been formatted and checked, the last step is to submit a pull request following the guidelines of `CONTRIBUTING.md`.
