---
title: "Extending ordr: FA"
output: 
  rmarkdown::html_vignette:
    toc: TRUE
vignette: >
  %\VignetteIndexEntry{Extending ordr: FA}
  %\VignetteEngine{knitr::rmarkdown}
  %\VignetteEncoding{UTF-8}
---

```{r, setup, include=FALSE}
knitr::opts_chunk$set(
  comment = '', fig.width = 6, fig.height = 6
)
```

# Introduction

This vignette demonstrates how to integrate methods for a new ordination class into **ordr.extra**. It follows the PCA vignette, which for reasons of underlying theory is a simpler case.
Since we'll refer to the prequel here occasionally, we recommend beginning there.

```{r, message = FALSE}
library(ordr)
library(psych)
```

## Factor Analysis

Factor analysis (FA) is a family of dimension-reducing methods of geometric data analysis. Unlike PCA, FA assumes the existence of latent variables (factors) behind a set of data, the number of which is less than the number of variables originally taken to be in the set. Like principal components, the factors in FA are linear combinations of the variables, and the variables are said to load onto the factors. However, factors are chosen to explain correlations between the variables, not to capture maximal variance; it is not assumed that factors are uncorrelated and orthogonal.

The core R distribution contains `stats::factanal()`, and **ordr** already recognizes its output. This vignette focuses on `psych::fa()`, with `factanal` methods as points of reference. Its integration into **ordr** will be much like that of `psych::principal()`, due to similarities in the linear algebra and in their interpretations, but several pitfalls must be avoided.

```{r, echo = TRUE, warning = FALSE, message = FALSE}
scaled_swiss <- scale(swiss)
fa_stats <- factanal(scaled_swiss, factors = 2, rotation = "none", scores = "regression")
(fa_tbl_ord <- as_tbl_ord(fa_stats))
```

## Factor Indeterminacy

FA assumes correlations between the factors, which leads to the issue of factor indeterminacy: Different optimization procedures may produce different sets of $k$ factors.
Moreover, the factors obtained are not ranked and may be inconsistent across different choices of $k$. This is in stark contrast to the fact that the first $l \leq k$ principal components are the same regardless of $k$.
Compare the results using $k=1$ and $k=2$:

```{r, echo = TRUE, warning = FALSE, message = FALSE}
fa(scaled_swiss, nfactors = 1)$loadings
fa(scaled_swiss, nfactors = 2)$loadings

op <- par(mfrow = c(1, 2))

fa(scaled_swiss, nfactors = 1) |>
  biplot.psych(main = "Biplot of FA with 1 factor")

fa(scaled_swiss, nfactors = 2) |>
  biplot.psych(main = "Biplot of FA with 2 factors")

par(op)
```

## Matrix Decomposition

Factor analysis is based on the following model:
$$R_{p\times{p}} = L_{p\times{k}}(L_{p\times{k}})^\top + U_{p\times{p}}^2$$
Note that FA uses the correlation matrix $R$ rather than the covariance matrix, which can also be obtained by centering and scaling the data:
$$(\bar{X}_{n\times{p}})^\top\bar{X}_{n\times{p}} = L_{p\times{k}}(L_{p\times{k}})^\top + U_{p\times{p}}^2$$
As with PCA via EVD, we decompose this matrix into the product of a loadings matrix $L$ and its transpose. However, there is no diagonal matrix of inertia here, and we add to the decomposition a residuals matrix $U^2$.
We can check this on the decomposition returned by `psych::fa()`:

```{r}
# correlation matrix
cor_swiss <- cov(scaled_swiss)
# factor analysis
fa_psych <- 
  psych::fa(r = swiss, nfactors = 2L, rotate = "none", scores = "regression", fm = "ml")
# factor analysis loadings and residuals
L <- fa_psych$loadings
Usquared <- fa_psych$residual
# equivalence
cor_swiss / (L %*% t(L) + Usquared)
```

As mentioned above, the loadings matrix $L$ depends on the FA method; the `fm` argument selects among 12 methods provided by **psych**.
In this vignette, we'll hold the method fixed as we expand **ordr** around this new class, then verify that the expansion accommodates other methods correctly.

# Recovery Methods

We'll first assemble the recoverers one by one.
Then we'll write and test the `as_tbl_ord()` method for the payoff.

## Row and Column Elements

As with `psych::principal()` we take the factor loadings as the active column elements and the scores as supplementary row elements.
An additional complication in FA, from a PCA perspective, is that several methods are available to compute case scores, based on different choices of weights. The default, regression scores, uses the weight matrix $W_{p\times{k}} = R^{-1}_{p\times{p}}L_{p\times{k}}$.
The scores are obtained from the centered data matrix by multiplication on the right, $S_{n\times{k}} = \bar{X}_{n\times{p}}W_{p\times{k}}$.

<details>
<summary> Check how scores are computed </summary>

```{r}
# the matrix of 1s indicates equality
(w <- fa_psych$weights) / (solve(cor_swiss) %*% fa_psych$loadings)
# the matrix of 1s indicates equality
head(fa_psych$scores / (scaled_swiss %*% w))
```

</details>

The scores do not necessarily multiply with the loadings to recover the data, but we may obtain a matrix decomposition that justifies their inclusion as row elements.
The weight matrix $W_{p\times k}$ is not invertible, but the _pseudoinverse_ $W^+_{k\times p} = ((W_{p\times k})^\top W_{p\times k})^{-1}(W_{p\times k})^\top$ is a left inverse in the sense that $W^+W=I_{k\times k}$ when $W$ is full-rank and has the nice property that $WW^+W = W$.
We then have $S_{n\times k}W^+_{k\times p} = \bar{X}_{n\times p}W_{p\times k}W^+_{k\times p}$---that is, the scores multiply with the pseudoinverse weights to recover an idempotent transformation of the scaled data matrix.

<details>
<summary> Check how the pseudoinverse multiplies with the weights </summary>

```{r}
# define pseudoinverse W^+
w_inv <- solve(t(w) %*% w) %*% t(w)
# left and right products with W
w_inv %*% w
w %*% w_inv
# idempotency of right product
( w %*% w_inv %*% w ) / w
```

</details>

This decomposition situates the scores as row elements and also prompts us to include (the transpose of) the pseudoinverse weight matrix as a set of supplementary column elements.
Indeed, $(W^+_{k \times p})^\top$ and $L_{p\times k}$ share the same dimensions, and we can reason that both contain full inertia.

<details>
<summary> How the same inertia is conferred on $(W^+)^\top$ as on $L$ </summary>

Let $d$ denote the inertia contained in $X$.
Then $(\bar{X}_{n\times p})^\top\bar{X}_{n\times p} = R_{p\times p}$ contains $d^2$. Recalling that $L_{p\times k}$ contains $d$, it follows that $W_{p\times{k}} = R^{-1}_{p\times{p}}L_{p\times{k}}$ contains $d^{-2}d = d^{-1}$.
Consequently, $W^+_{k\times p} = ((W_{p\times k})^\top W_{p\times k})^{-1}(W_{p\times k})^\top$ contains $(d^{-2})^{-1}d^{-1} = d^2d^{-1} = d$.
This is also consistent with the scores being in standard coordinates, as determined for `psych::principal()`.

</details>

We can now define the row and column recoverers, largely mimicking those of the `principal` class:

```{r}
# no active row elements
recover_rows.fa <- function(x) {
  matrix(
    nrow = 0, ncol = ncol(x[["loadings"]]),
    dimnames = list(NULL, colnames(x[["loadings"]]))
  )
}
# loadings as active column elements
recover_cols.fa <- function(x) {
  unclass(x[["loadings"]])
}
# scores as supplementary row elements
recover_supp_rows.fa <- function(x) {
  if (is.null(x[["scores"]])) {
    matrix(numeric(0), nrow = 0, ncol = 0)
  }
  else
    x[["scores"]]
}
# transpose pseudoinverse weight matrix as supplementary column elements
recover_supp_cols.fa <- function(x) {
  t( solve(t(x[["weights"]]) %*% x[["weights"]]) %*% t(x[["weights"]]) )
}
# illustrations
recover_cols(fa_psych)
recover_rows(fa_psych)
recover_supp_cols(fa_psych)
head(recover_supp_rows(fa_psych))
```

<details>
<summary> Check that elements agree with those from `factanal` </summary>

Let us benchmark these choices against our `factanal` object from earlier.
While `stats::factanal()` returns no supplementary columns, the remaining three sets of elements are approximately equal:

```{r}
recover_cols(fa_stats) / recover_cols(fa_psych)
recover_rows(fa_stats) / recover_rows(fa_psych)
head(recover_supp_rows(fa_stats) / recover_supp_rows(fa_psych))
```

</details>

### Conference of Inertia

Because FA permits correlations between factors, the procedure does not produce a matrix of inertias; the variance cannot be partitioned and attributed to the factors as it is to the components in PCA.
Whereas the first $l \leq k$ principal components always correspond to the $l$th eigenspace, only the $k$th eigenspace is expressible as the span of the $k$ factors.

In order to accommodate the SVD-based recoverers, we must extract values analogous to inertia that are meaningful in the FA setting.
In PCA, the covariance matrix is decomposed as $X^\top X = (V D) (D V^\top)$, and the inertia of each principal component---each column of $V D$---is the sum of the squares of its principal loadings, multiplied by $n-1$ to account for standardization:

<details>
<summary> Recover inertia from columns in PCA </summary>

```{r, eval = FALSE}
recover_inertia.principal(pca_psych)
colSums(pca_psych$loadings ^ 2) * (nrow(scaled_iris) - 1)
```

</details>

For FA, we borrow this calculation and compute the variances of the factors in the same way, with $L$ in place of $V D$.
This means that full inertia is conferred onto $L$.
Since the inner products of the scores and loadings approximate the data matrix, $X \approx S L$, the scores must have no inertia conferred.
It follows that `psych::fa()` has a conference of $(0,1)$.

<details>
<summary> Check that $S L$ approximates $X$ </summary>

First, we show that the scores multiplied by the transposed loadings gives an approximation of the data matrix.

```{r}
# create "fa" object, now with six factors
fa_psych6 <- psych::fa(swiss, nfactors = 6, rotate = "none")

stdized_approx <- fa_psych6$scores %*% t(as.matrix(fa_psych6$loadings))
means <- colMeans(swiss)
sds <- apply(swiss, 2, sd)
swiss_approx <- sweep(stdized_approx, 2, sds, "*") # de-scale
swiss_approx <- sweep(swiss_approx, 2, means, "+") # de-center

# values close to 1 indicate near equality
head(swiss_approx / swiss)
```

</details>

It also means that we focus on the geometry of the shared factors alone, ignoring the unique factors, which don't contribute to the loadings.
Recall that, in FA, the total variance is the sum of _shared variance_ and _uniquenesses_.
The next fold shows that our "inertia" from the loadings can be added to the uniquenesses to recover the total.

<details>
<summary> Recompose inertia with uniqueness to obtain total variance </summary>

Note below that the total is rescaled by $n-1$, bringing it to $n$, or $1$ for each variable:

```{r}
# inertia from loadings
common_var <- (colSums(fa_psych$loadings^2))
# uniquenesses
uniq <- fa_psych$uniquenesses
# shared variance + uniquenesses
sum(common_var) + sum(uniq)
```

</details>

These observations lead us to our recoverers for conference and for inertia:

```{r}
recover_conference.fa <- function(x) {
  c(0, 1)
}

recover_inertia.fa <- function(x) {
  colSums(x[["loadings"]] ^ 2)
}

recover_conference.fa(fa_psych)
recover_inertia.fa(fa_psych)
```

<details>
<summary> Check that elements agree with those from `factanal` </summary>

Finally, we check that these values basically agree with those recovered from our `factanal` object:

```{r}
recover_inertia(fa_stats)
```

</details>

### Augmentation

```{r}
library(tibble)
```

The unique factors, or uniquenesses, are a property of the manifest variables that we haven't given a clear geometric interpretation, so the place to provide them to the user is as augmentation. The `fa` object returned by `psych::fa()` also contains `communality` and `complexity` values, so we pool all three into the column augmentation.
The coordinate and row augmentors are perfunctory: names and element types only:

```{r}
recover_coord.fa <- function(x) {
  colnames(x[["loadings"]])
}

recover_aug_coord.fa <- function(x) {
  tibble(
    name = ordr:::factor_coord(recover_coord(x))
  )
}

recover_aug_rows.fa <- function(x) {
  res <- tibble(.rows = 0L)
  
  # scores as supplementary points
  if (!is.null(x[["scores"]])) {
    name <- rownames(x[["scores"]])
    res_sup <- if (is.null(name)) {
      tibble(.rows = nrow(x[["scores"]]))
    } else {
      tibble(name = name)
    }
  } else {
    res_sup <- tibble(.rows = 0L)
  }
  
  # supplement flag
  res$.element <- "active"
  res_sup$.element <- "score"
  as_tibble(dplyr::bind_rows(res, res_sup))
}

recover_aug_cols.fa <- function(x) {
  name <- rownames(x[["loadings"]])
  res <- if (is.null(name)) {
    tibble(.rows = nrow(x[["loadings"]]))
  } else {
    tibble(name = name)
  }
  res$uniqueness <- x$uniquenesses
  res$communality <- x$communality
  res$complexity <- x$complexity
  
  # supplement flag
  res$.element <- "active"
  # reorder columns
  res <- res[c(".element", setdiff(names(res), ".element"))]
  
  # transposed pseudoinverse of weights as supplementary points
  name <- rownames(x[["weights"]])
  res_sup <- if (is.null(name)) {
    tibble(.rows = nrow(x[["weights"]]))
  } else {
    tibble(name = name)
  }
  
  # supplement flag
  res_sup$.element <- "inv_weight"
  # reorder columns
  res_sup <- res_sup[c(".element", setdiff(names(res_sup), ".element"))]
  as_tibble(dplyr::bind_rows(res, res_sup))
}
```

<details>
<summary> Check that the augmentors have correct numbers of rows </summary>

To preempt bugs (and possibly confusing error messages), we verify that the augmentors have the expected number of rows:

```{r}
# should yield two rows for the two factors
recover_aug_coord.fa(fa_psych) |> nrow()

# should yield 47 rows for the 47 rows in the data frame/scores
recover_aug_rows.fa(fa_psych) |> nrow()

# should yield twelve rows for the six rows of weights and six rows of loadings
recover_aug_cols.fa(fa_psych) |> nrow()
```

</details>

See the `principal` vignette for more details.

We can now assemble the `tbl_ord` object by writing a method for `as_tbl_ord()`:

```{r}
as_tbl_ord.fa <- getFromNamespace("as_tbl_ord_default", "ordr")

fa_psych |> 
  as_tbl_ord() |>
  augment_ord()
```

### Examples

For our examples, to be put in a file `ex-methods-psych-fa-<data>.r`, we replicate those of `ex-methods-factanal-swiss.r`:

```{r}
# data frame of Swiss fertility and socioeconomic indicators
class(swiss)
head(swiss)
# perform factor analysis
swiss_fa <- psych::fa(
  r = swiss,
  nfactors = 2L, rotate = "varimax", scores = "regression", fm = "ml"
)

# wrap as a 'tbl_ord' object
( swiss_fa <- as_tbl_ord(swiss_fa) )

# recover loadings
get_cols(swiss_fa, elements = "active")
# recover scores
head(get_rows(swiss_fa, elements = "score"))

# augment column loadings with uniquenesses
(swiss_fa <- augment_ord(swiss_fa))
```

### Unit Tests

Similarly, we adapt the tests from `test-stats-factanal.r` for the unit test script. Note that, in **ordr.extra**, unit tests for GDA methods of the same package are included in a single script.

```{r}
library(testthat)

fit_fa <- psych::fa(
  r = swiss,
  nfactors = 2L, rotate = "varimax", scores = "regression", fm = "ml"
)

test_that("'fa' accessors have consistent dimensions", {
  expect_equal(ncol(get_rows(fit_fa)), ncol(get_cols(fit_fa)))
  expect_equal(ncol(get_rows(fit_fa)), length(recover_inertia(fit_fa)))
})

test_that("'fa' has specified distribution of inertia", {
  expect_type(recover_conference(fit_fa), "double")
  expect_vector(recover_conference(fit_fa), size = 2L)
})

test_that("'fa' augmentations are consistent with '.element' column", {
  expect_equal(
    ".element" %in% names(recover_aug_rows(fit_fa)),
    ".element" %in% names(recover_aug_cols(fit_fa))
  )
})

test_that("`as_tbl_ord()` coerces 'fa' objects", {
  expect_true(valid_tbl_ord(as_tbl_ord(fit_fa)))
})
```

### Submission

As with `psych::principal()` , we should conclude by running `devtools::check()`, then submit a pull request following the guidelines of `CONTRIBUTING.md`.
