---
title: "Extending ordr: PCA"
output: 
  rmarkdown::html_vignette:
    toc: TRUE
vignette: >
  %\VignetteIndexEntry{Extending ordr: PCA}
  %\VignetteEngine{knitr::rmarkdown}
  %\VignetteEncoding{UTF-8}
---

```{r, setup, include=FALSE}
knitr::opts_chunk$set(
  comment = '', fig.width = 6, fig.height = 6
)
```

# Introduction

This is the first of 3 vignettes that detail the process of extending **ordr** to incorporate new ordination methods. This vignette focuses on principal component analysis, a straightforward SVD-based technique, while the sequel focuses on factor analysis, a much more complicated case. We demonstrate how to extract key elements from each class based on a shared underlying mathematical theory and to make these available to **ordr** using a set of methods based on this theory. We also illustrate how unit tests and examples are written to complete the contribution. The final vignette applies both contributions to a common data set to showcase their commonalities and differences. The vignettes are intended for R users with experience in geometric data analysis and with the tidyverse.

If an extension of **ordr** to a new ordination class is agreed upon, the first step will be to open an issue that includes the checklist from this developer vignette: `vignette("extension-checklist")`.

```{r, message=FALSE}
library(ordr)
library(psych)
```

## Principal Component Analysis

Principal component analysis (PCA) is a geometric statistical method by which multivariate data is reduced to fewer dimensions, expending as little variance as possible in the process---that is, the resulting dimensions, called _principal axes_, contain as much of the variance as possible. The goal is to more easily visualize and interpret patterns in the data. PCA is useful in various fields; this vignette will conclude with an example using psychometric data.

## EVD and SVD

PCA can be done by applying either of two matrix decomposition methods to a centered and (often) scaled data matrix $\bar{X}_{n\times p}$.

One method, used by `stats::prcomp()`, is singular value decomposition (SVD).
Assuming $n > p$, SVD obtains the decomposition
$$\bar{X}_{n\times{p}} = U_{n\times{p}}D_{p\times{p}}(V_{p\times{p}})^\top\text{,}$$
where the column vectors of $U$ and of $V$ have length $1$ and are pairwise perpendicular and $D$ is diagonal.
The columns of $U$ are called the _left singular vectors_ of $\bar{X}$ and those of $V$ the _right singular vectors_; the diagonal elements of $D$ are the _singular values_.
Based on the equation, elements from $U$ and from $V$, respectively, are referred to as "row" and "column" elements.

```{r}
scaled_iris <- scale(iris[1:4])
pca_stats <- prcomp(scaled_iris, rank. = 4)
```

The other method uses eigenvalue decomposition (EVD), or just eigendecomposition, as in `stats::princomp()` and in `psych::principal()`. The EVD of a full-rank square matrix $M$ is the unique decomposition
$$M = V \Lambda V^\top$$
with $\Lambda$ diagonal; the entries of $\Lambda$ are called the eigenvalues, and the columns of $V$ are called the eigenvectors, of $M$.
For PCA, we obtain $U$ as the matrix of eigenvectors of $\bar{X}\bar{X}^\top$ and $V$ as the that of $\bar{X}^\top\bar{X}$. The positive eigenvalues from these two decompositions agree, and their square roots are the singular values of $\bar{X}$. The fold demonstrates these connections between SVD and EVD.

```{r}
( pca_psych <- principal(scaled_iris, nfactors = 4, rotate = "none") )
```

<details>
<summary> Derive EVD from SVD </summary>

To understand EVD in terms of SVD, again take $X=UDV^\top$. Then when we consider the covariance matrix $\bar{X}^\top \bar{X}$, we have

$$
\begin{align*}
(\bar{X}_{n\times{p}})^\top \bar{X}_{n\times{p}} &= (U_{n\times{p}}D_{p\times{p}}(V_{p\times{p}})^\top )^\top  U_{n\times{p}}D_{p\times{p}}(V_{p\times{p}})^\top \tag{1}\\
&= ((V_{p\times{p}})^\top )^\top D_{p\times{p}}^\top (U_{n\times{p}})^\top U_{n\times{p}}D_{p\times{p}}(V_{p\times{p}})^\top \tag{2}\\
&= V_{p\times{p}}D_{p\times{p}}D_{p\times{p}}(V_{p\times{p}})^\top \tag{3}\\
&= V_{p\times{p}}(D_{p\times{p}})^2(V_{p\times{p}})^\top \tag{4}.
\end{align*}
$$

Note: we use the orthogonality of $U$ to obtain (3) as $U^\top = U^{-1}$.

Indeed, the singular values of $\bar{X}$ are the square roots of the eigenvalues of $\bar{X}^\top\bar{X}$, meaning $D^2 = \Lambda$.

</details>

In reference to factor analysis, $V$ is often called the loadings matrix---the columns contain the linear coefficients of the measured variables as they load onto each principal component---while $U$ contains the scores.
<!--In EVD, $\bar{X}$ must be multiplied into the decomposition to obtain scores, demonstrated later.-->
Depending on whether the investigator's focus is on the variables or the cases, they may want to study and visualize the loadings or the scores, respectively. When focusing on the scores, the row data (the cases) may be studied using _principal coordinates_, the rows of $UD$, while the column data (the variables) are left in _standard coordinates_, the rows of $V$. Here, the inertia of $D$ has been _conferred_ onto the left singular vectors, changing the space in which they are being coordinatized.
When focusing instead on the variables, the inertia will often be conferred onto the right singular vectors, putting the variables in principal coordinates (the rows of $VD$) and the cases in standard coordinates (the rows of $U$).

# Recovery Methods

By *recovery methods* or *recoverers* we mean the functions that retrieve key elements from an ordination object. The essential recoverers extract the components of the decomposition just discussed, while others extract supplementary elements and annotations specific to the technique and its implementation.

While these recovery methods will depend on the GDA technique, the helper function `ordr::as_tbl_ord()` reduces redundancy and expedites debugging, so we give it a method for the `principal` class:

```{r}
as_tbl_ord.principal <- getFromNamespace("as_tbl_ord_default", "ordr")
```

This method draws from the recoverers to produce a `tbl_ord`, illustrated below with the pre-loaded set of Anderson's iris data:

```{r}
( tbl_ord <- as_tbl_ord(pca_stats) )
```

## Row and Column Elements

We will begin by defining the following recoverers for a `principal` object: 

  * `ordr::recover_cols()` returns $V$ or some $VD^b$, which factors into the SVD from the right.
  * `ordr::recover_rows()` returns $U$, or more generally $UD^a$ for some $a$, which factors in from the left.

In an SVD $\bar{X} = UDV^\top$, the left and right matrix factors contain coefficients of the row elements and column elements, respectively, of the original data $X$ in a new coordinate space. This is why we refer to elements of the left factor $U$ (or $UD$) as "row elements" and those of the right factor $V$ (or $VD$) as "column elements". With SVD, both factors are said to contain "active" elements.

An EVD-based method might be based on the row inner products $X X^\top$ or on the column inner products $\bar{X}^\top \bar{X}$. Either way, the EVD yields a symmetric factorization $V \Lambda V^\top = (VD)(VD)^\top$; the "left" and "right" factors are the same.
EVD-based PCA uses the latter, a decomposition of the covariance matrix $X^\top X$ (which happily means that the $V$ in this paragraph is the same as that in the last). We therefore interpret the loadings $VD$ as active column elements, and we write the column recovery method to extract this matrix:

```{r}
recover_cols.principal <- function(x) {
  unclass(x[["loadings"]])
}
recover_cols.principal(pca_psych)
```

We obtain no active row elements from this class, so we write the row recovery method to return a matrix with no rows (elements) but the correct number and names of the columns (artificial coordinates):

```{r}
recover_rows.principal <- function(x) {
  matrix(
    nrow = 0, ncol = ncol(x[["loadings"]]),
    dimnames = list(NULL, colnames(x[["loadings"]]))
  )
}
recover_rows.principal(pca_psych)
```

We do obtain case scores, which are unambiguously row elements; however, they are not obtained through SVD but computed internally. We can still verify that they multiply with the loadings to reconstruct $\bar{X}$:

<details>
<summary> Check that scores and loadings reconstruct data </summary>

We divide the scores--loadings product elementwise by the data matrix and print the first few rows:

```{r}
head(
  (pca_psych$scores %*% t(pca_psych$loadings)) / scaled_iris
)
```

We use element-wise division to show equality rather than `all.equal()` or `identical()` because rounding errors may result in small differences in computation despite mathematical equivalence.
The matrix of all 1s indicates that the original matrices were equal.

</details>

Hence, the scores are supplementary row elements, and we can go ahead and write a recovery method for them. We must also account for cases where `scores = FALSE` is passed, resulting in no `scores` element in the `principal`-class output.

```{r}
recover_supp_rows.principal <- function(x) {
  if (is.null(x[["scores"]])) {
    tibble(numeric(0), nrow = 0, ncol = ncol(x[["loadings"]]))
  }
  else {
    x[["scores"]]
  }
}
head(recover_supp_rows.principal(pca_psych))
```

## Inertia and its Conference

  * `ordr::recover_inertia()` returns the eigenvalues of our EVD, or equivalently the squared singular values of our SVD---the diagonal entries of $D^2$.

From the documentation, `psych::principal()` returns the eigenvalues in the `values` element of the output:

```{r}
pca_psych$values
```

We can validate these values against those returned using the recovery method for the `prcomp` class:

```{r}
recover_inertia(pca_stats) / pca_psych$values
```

Notice that $149=n-1$, indicating that these eigenvalues have been scaled by the degrees of freedom. To obtain the inertia, we need then to multiply them by $n-1$:

```{r}
recover_inertia.principal <- function(x) {
  x[["values"]] * (nrow(x[["scores"]]) - 1L)
}
recover_inertia.principal(pca_psych)
```

This method is consistent with that for the `prcomp` class.

These values are usually given the names of the artificial coordinates. Because these names are used for several other purposes, they have a separate recovery method. For this, we define 

  * `ordr::recover_coord()` to return the names of the coordinate axes of our ordination object.

As with the row element recoverers above, we obtain these names from the column names of the loading matrix:

```{r}
recover_coord.principal <- function(x) {
  colnames(x[["loadings"]])
}
recover_coord.principal(pca_psych)
```

The inertia is established, but the methods must also account for how it is conferred on the matrix factors, which brings us to the next recoverer.

  * `ordr::recover_conference()` indicates how the inertia is distributed to the rows or columns ($U$ or $V$) in the object. (**stats** and **psych** make different choices here.)

<details>
<summary> Implications of different conferences of inertia </summary>

The "conference" governs the geometry of biplots and matters crucially to their interpretation, but the matrix factors returned by different techniques may confer it differently. To avoid having to manipulate the matrix factors during recovery, we use the `recover_conference()` method to inform **ordr** how this conference is done.

Not distributing the inertia (i.e. not multiplying out the matrix $D$ in SVD or EVD) maintains $U$ and $V$ as matrices of unit vectors. Thus, we say that the cases and variables are in standard coordinates. But when we _do_ distribute the inertia to one or both other matrices in the decomposition, we obtain the cases and/or variables in principal coordinates.

Assuming we desire principal coordinates, EVD-based techniques leave only the option of conferring inertia onto one matrix of singular vectors. This depends on the implementation (and how we interpret it), and in the case of `psych::principal()` it would be the right singular vectors. However, the conference of inertia in SVD-based techniques depends on the user's motivation. In short, if the user wishes to maintain the approximate Euclidean distances between data cases, then conference onto the left singular vectors is ideal. If instead the user wishes to maintain the approximate correlations between variables, then conference onto the right singular vectors is ideal.

```{r}
recover_conference(pca_stats)
```

The conference recovery method is, so far, always a static vector, since the methods don't confer inertia differently based on the input. Here, we see that `stats::prcomp()` automatically confers all inertia onto the left singular vectors (a conference of $(0,1)$, alternatively, would indicate that a PCA has conferred its inertia onto the right singular vectors). This means that the scores of `pca_stats` are in principal coordinates and the loadings are in standard coordinates.

</details>

This is straightforward for the loadings: We expect from the EVD that they are in principal coordinates, i.e. that they have full inertia, and we checked that the loadings matrix agrees with the full-inertia loadings obtained from `stats::prcomp()`. So, $L = VD$. We also checked that the product $SL^\top$ reconstructs $\bar{X}$, which suggests that $S = U$. So, the scores should be in standard coordinates, i.e. they have been conferred no inertia. Symbolically:

$$ X = UDV^\top = (UD^0)(VD^1)^\top = SL^\top$$

This yields a conference of $(0,1)$. A careful check follows.

<details>
<summary> Check that the loadings matrix has full inertia </summary>

```{r}
# X'X from EVD
cov <- cov(scaled_iris)
evd <- eigen(cov)
# observe loadings matrix equals VD up to sign
( evd$vectors %*% diag(sqrt(evd$values))) / unclass(pca_psych$loadings )
```

Hence we have $L_{p\times{p}} = V_{p\times{p}}D_{p\times{p}}$.

</details>

As such, we define the following recovery method:

```{r}
recover_conference.principal <- function(x) {
  c(0, 1)
}
recover_conference.principal(pca_psych)
```

## PCA Followed by Rotation

In factor analysis, the loadings---a transformation of the measured or "manifest" variables into a smaller set of latent variables---will often be composed with a rotation---a change of basis to a new but same-sized set of latent variables---to achieve "simple structure" in which the latent variables are easier to interpret. We'll discuss this in more detail in the FA vignette. Because PCA has long been appropriated for FA, such rotations may also be applied to the results of PCA, and this functionality is provided by **psych**.

Rotations may be either orthogonal or oblique. We compute and illustrate both in the next two code chunks, using an oblique rotation provided by **GPArotation**.

<details>
<summary> Remark about precision of language </summary>

Mathematically, oblique rotations are not proper rotations, which are angle-preserving. Following the FA literature, we refer to both as "rotations" but always qualify those that are oblique.

Fundamental to PCA are the orthogonality of the principal components and the minimality of the residual variance of each principal subspace. Oblique rotations compromise the former, while even orthogonal rotations undermine the latter. For these reasons, it is contested whether PCA with rotation should still be considered PCA. For clarity, we refer to these models as "PCA _followed by_ rotation".

</details>

```{r, message=FALSE}
library(GPArotation)
( pca_ortho <- principal(scaled_iris, rotate = "varimax", nfactors = 4) )
( pca_obliq <- principal(scaled_iris, rotate = "oblimin", nfactors = 4) )
```

```{r, warning=FALSE, fig.width=8, fig.height=3}
op <- par(mfrow = c(1, 3))
biplot.psych(pca_psych, choose = c(1, 2), main = "no rotation")
biplot.psych(pca_ortho, choose = c("RC1","RC2"), main = "orthogonal rotation")
biplot.psych(pca_obliq, choose = c("TC1","TC2"), main = "oblique rotation")
par(op)
```

The biplots suggest that the rotations twist the manifest variables (vectors) to better align with the latent variables (axes); in fact, they twist the latent variables to better align with the manifest variables. Because all 4 factors (principal components) were obtained, the rotations are not constrained to two dimensions, and the geometry of the vectors in the 2-dimensional projection warps. However, both rotations preserve the correlations between the manifest variables in the transformed space. (The angles between vectors appear to change dramatically under the oblique rotation, but recall that the rotated principal axes are no longer orthogonal; the biplots do not communicate this and even obscure it.)

<details>
<summary> Check and interpret rotated loadings </summary>

This extended fold examines the linear algebra at play in each rotation. In PCA followed by an orthogonal rotation, the rotated loadings are the original loadings matrix $L_{p\times{p}}$ multiplied by the orthogonal rotation matrix $T_{p\times{p}}$:

```{r}
pca_ortho$loadings[, c("RC1", "RC2", "RC3", "RC4")] /
  (pca_psych$loadings %*% pca_ortho$rot.mat)
```

In the oblique case, we obtain two loading matrices. The first is the _pattern matrix_ $L_{p\times{p}} T_{p\times{p}}$. The `$loadings` element extractable from an oblique-transformed `principal` object is the pattern matrix:

```{r}
pca_obliq$loadings[, c("TC1", "TC2", "TC3", "TC4")] /
  (pca_psych$loadings %*% pca_obliq$rot.mat)
```

The pattern matrix quantifies how much each principal component loads into each variable. 

The second loadings object is the _structure matrix_ $S_{p\times{p}} = L_{p\times{p}}T_{p\times{p}}\Phi_{p\times p}$. The structure matrix is obtained from the pattern matrix via multiplying by the interfactor correlation matrix $\Phi_{p\times p}$.

```{r}
(pca_obliq$loadings %*% pca_obliq$Phi) /
  pca_obliq$Structure
```
The structure matrix contains correlations between factors and variables, and it tends to be more stable between samples.

In the orthogonal case, $\Phi_{p\times p}$ is the identity matrix, which is why we have just one loadings matrix. Accordingly, a `principal` object only contains `$Structure` and `$Phi` elements if it contains an oblique rotation.

The choice between an orthogonal and oblique rotation depends on our priorities. An orthogonal rotation keeps the model more faithful to the data. But if we want to convey significant correlation in the variables or to more easily interpret a biplot, then an oblique rotation may be more appropriate. Various different rotations suitable for different goals exist within either category.

</details>

One way **ordr** might organize this information would be to separately recover the unrotated loadings and the rotation matrix. This would require some transformation, since the `principal` class stores the already-rotated loadings, but it might be worth the trouble to disentangle other aspects of the class and to generalize the recovery process to more methods. On the other hand, it might be misguided to recover the rotation matrix as an element, as it would enable questionable practices like overlaying un-rotated loadings with scores based on the rotated loadings. For now, **ordr** recovers elements for the rotated model:

```{r}
recover_cols.principal(pca_obliq)
recover_cols.principal(pca_ortho)
```

Either way, the methods must be informed by the additional caution and checks above.

## Augmentation

  * `ordr::recover_aug_rows()` and `ordr::recover_aug_cols()` do not recover raw elements of the PCA but rather reassemble additional, often non-numeric elements that augment the row and column coordinates.

We have recovered all of the **ordr**-relevant geometric data from the `principal` object. What remains are any pieces of information in the class that annotate the matrix factors. This information is organized in each case as a tibble, not a matrix. For convenience, we attach the **tibble** package:

```{r}
library(tibble)
```

The most important annotations are the names, which will usually have been taken from the row and column names of the data set $\bar{X}$. But they may also include information specific to these units (cases and variables for PCA) that was carried over from the data object, e.g. the `Species` column from the `iris` data, or that governed pre-processing, e.g. the variable means and standard deviations when variables are scaled before PCA.

The row and column augmentation recoverers must also comport with the active and any supplementary elements recovered by those methods. For example, our row augmentation function must recognize that no active row elements are recovered and that the supplementary elements, if available for recovery, correspond to the cases:

```{r}
recover_aug_rows.principal <- function(x) {
  # no active elements
  res <- tibble(.rows = 0L)
  
  # scores as supplementary elements
  if (!is.null(x[["scores"]])) {
    name <- rownames(x[["scores"]])
    res_sup <- if (is.null(name)) {
      tibble(.rows = nrow(x[["scores"]]))
    } else {
      tibble(name = name)
    }
  } else {
    res_sup <- tibble(.rows = 0L)
  }
  
  # supplement flag
  res$.element <- "active"
  res_sup$.element <- "score"
  as_tibble(dplyr::bind_rows(res, res_sup))
}
```

Our column augmentation function concerns only active elements. In addition to the variable names, we augment the column elements with the means and standard deviations obtained during pre-processing and included in the `principal` object:

```{r}
recover_aug_cols.principal <- function(x) {
  name <- rownames(x[["loadings"]])
  res <- if (is.null(name)) {
    tibble(.rows = nrow(x[["loadings"]]))
  } else {
    tibble(name = name)
  }
  res$.element <- "active"
  res
}
```

These values are required for some biplot layers like calibrated axes.

Finally, the coordinate augmentation function provides the coordinate names and their standard deviations, which are provided in the `principal` object:

```{r}
recover_aug_coord.principal <- function(x) {
  data.frame(
    name = recover_coord.principal(x),
    sdev = sqrt(x[["values"]][seq(1, ncol(x[["loadings"]]))])
  )
}
```

<details>
<summary> Check numbers of rows </summary>

To preempt bugs, we should ensure that the number of rows in each augmented data frame agree with those of the corresponding extracted matrix.

```{r}
nrow(recover_aug_rows.principal(pca_psych)) ==
  nrow(recover_rows.principal(pca_psych)) +
  nrow(recover_supp_rows.principal(pca_psych))
nrow(recover_aug_cols.principal(pca_psych)) ==
  nrow(recover_cols.principal(pca_psych))
nrow(recover_aug_coord.principal(pca_psych)) ==
  length(recover_inertia(pca_psych))
```

</details>

We have now defined the methods required to obtain the `tbl_ord` wrapper for a `principal` object:

```{r}
pca_psych |> 
  as_tbl_ord() |>
  augment_ord()
```

This can also now be done "under the hood" by `ordinate()`:

```{r}
ordinate(scaled_iris, ~ principal(., nfactors = 4, rotate = "none"))
```

# Package Infrastructure

With the recovery methods written and checked, we round out our contribution with unit tests, examples, and written documentation.

## Unit Tests

First, we draw from our lessons above to write a script of unit tests for these recoverers. We locate the script in `tests/testthat/` and name it following the pattern `test-<package>-<class>.r`. To demonstrate that our methods pass the tests, we attach the **testthat** package:

```{r}
library(testthat)
```

We write the following tests, which will only be run if **psych** is installed:

```{r}
skip_if_not_installed("psych")

fit_principal <- psych::principal(iris[, -5], nfactors = 4, rotate = "none")

test_that("'principal' accessors have consistent dimensions", {
  expect_equal(ncol(get_rows(fit_principal)), ncol(get_cols(fit_principal)))
  expect_equal(ncol(get_rows(fit_principal)),
               length(recover_inertia(fit_principal)))
})

test_that("'principal' has specified distribution of inertia", {
  expect_type(recover_conference(fit_principal), "double")
  expect_vector(recover_conference(fit_principal), size = 2L)
})

test_that("'principal' augmentations are consistent with '.element' column", {
  expect_equal(".element" %in% names(recover_aug_rows(fit_principal)),
               ".element" %in% names(recover_aug_cols(fit_principal)))
})

test_that("`as_tbl_ord()` coerces 'principal' objects", {
  expect_true(valid_tbl_ord(as_tbl_ord(fit_principal)))
})
```

## Examples

We next adapt an examples script from those for other methods, but reflecting some of the special cases we encountered above. The file name  `ex-methods-psych-principal-iris.r` follows a similar convention but with the illustrative data set specified (allowing for more than one script if helpful), and the script also begins by conditioning itself on **psych** being installed:

```{r, eval=FALSE}
if (require(psych)) {# {psych}

# data frame of Anderson iris species measurements
class(iris)
head(iris)

# compute unscaled row-principal components of scaled measurements
iris[, -5] |>
  psych::principal(nfactors = 4, rotate = "none") |>
  as_tbl_ord() |>
  print() -> iris_pca

# compute varimax-rotated row-principal components of scaled measurements
iris[, -5] |>
  psych::principal(nfactors = 4, rotate = "varimax") |>
  as_tbl_ord() |>
  print() -> iris_pca_varimax

# recover observation principal coordinates
head(get_rows(iris_pca))

# recover rotated loadings
get_cols(iris_pca_varimax)

# augment measurement coordinates with names and scaling parameters
( iris_pca <- augment_ord(iris_pca) )
( iris_pca_varimax <- augment_ord(iris_pca_varimax) )

}# {psych}
```

## Documentation

Finally, we collect and document the recoverers in a file named `methods-psych-principal.r`, following the pattern `methods-<package>-<class>.r`:

```{r, eval=FALSE}
#' @title Functionality for principal components analysis ('principal') objects
#'
#' @description These methods extract data from, and attribute new data to,
#'   objects of class `"principal"` as returned by [psych::principal()].
#'
#'@details
#' 
#' <truncated>
#'
#' @name methods-principal
#' @author John Gracey
#' @template param-methods
#' @template return-methods
#' @family methods for eigen-decomposition-based techniques
#' @family models from the psych package
#' @example inst/examples/ex-methods-psych-principal-iris.R
NULL
```

The remainder of the file defines each recovery method, in each case preceded by tags that anchor it to the manual entry `methods-principal` and export it to the user. Note that the **roxygen2** tag `@family` cross-links this entry with related entries (EVD-powered and from the same package), while the tag `@example` calls the installed example script to include its code in the "Examples" section of the entry.

## Submission

The function `devtools::check()` is useful for verifying that the contribution will not generate errors or warnings, many of which won't show up when experimenting with code on the command line or an IDE like RStudio. It also first runs `devtools::document()` to "roxygenize" the documentation into an entry stored in the `man/` folder and incorporated into the package manual. Once the extension has been formatted and checked, the last steps are to edit `DESCRIPTION` accordingly, to write an entry in `NEWS.md` and to submit a pull request following the guidelines of `CONTRIBUTING.md`.
