---
title: "New Ordination Classes--Comparison"
author: "John Gracey"
date: "2025-10-02"
output: html_document
---

# Introduction

This vignette proceeds the two "Writing Recovery Methods for New Ordination Classes" vignettes, using the **ordr.extra** incorporations of `psych::principal()` and `psych::fa()` to illustrate some of the similarities and differences between methods at both the theoretical and practical level.

# Comparison of PCA and FA

```{r, echo = TRUE}
library(ordr.extra) # load {ordr.extra} to access `hcw` data below

scaled_hcw <- scale(hcw[, tail(colnames(hcw), 14)])
```

Having demonstrated PCA and FA independently of each other, we now apply both to one data set for comparison. The data comes from a 2022 study on burnout in healthcare workers (HCW) during the Covid-19 pandemic (Guastello et al., 2022). In particular, we use data from a fourteen-item Likert scale questionnaire, the Healthcare System Communication Questionnaire, surveying HCW perceptions of hospital administration. The original study applied PCA followed by an oblimin rotation (a type of oblique rotation) to the data to examine whether it had psychometric properties. It was seen that with two principal components, the questions generally loaded strongly onto just one of the two components. Those that loaded strongly onto the first principal component generally pertained to consideration from leadership, and those that loaded strongly onto the second generally pertained to structure in the hospital. The principal components were named accordingly.

We begin by recreating this PCA.

```{r}
# conduct PCA
hcw_pca <- scaled_hcw |>
  ordinate(~ principal(., nfactors = 2L, rotate = "oblimin"))
sum(hcw_pca$values[1:2]) / sum(hcw_pca$values[1:14]) # agrees with % variance captured from the paper
hcw_pca$loadings
```

These loadings approximately equal those in Guastello et al., 2022.

Next, we perform FA on the same centered and scaled data, again followed by an oblimin rotation. We also produce biplots for both analyses.

```{r}
# conduct FA
hcw_fa <- scaled_hcw |>
  ordinate(~ fa(., nfactors = 2L, rotate = "oblimin"))

biplot.psych(hcw_pca)
biplot.psych(hcw_fa)
```

The biplots are noticeably similar in both the geometry of the variable vectors and the distribution of observations. Comparing the loadings and scores from both GDA techniques, we can see that most results do align quite closely between techniques (we divide the matrices entry-wise so that values closer to 1 below indicate near-identical entries).

```{r}
unclass(hcw_fa$loadings / hcw_pca$loadings) # values close to 1 indicate near equality
head(hcw_fa$scores / hcw_pca$scores) # values close to 1 indicate near equality
```

This would suggest that most of the variance in the data is common and not unique to one variable. In both biplots, the variable vectors are mostly either close to horizontal or close to vertical, indicating strong correlations with those nearby variable vectors _and_ that most questions (i.e. variables) load strongly onto exactly one component. In the loadings matrix below, we see the latter is generally true.

```{r}
unclass(hcw_pca$loadings)
```

However, the previous biplots are not identical. There is an apparent elliptical shape to the data in our FA biplot (recall that the goal of FA is to unveil a "simple structure"). The PCA biplot appears to have slightly more spread due to PCA's focus on capturing total variance, not strictly shared variance as FA does. To see this effect at a slightly greater degree, we can consider the corresponding PCA with no rotation.

```{r}
# unrotated PCA
hcw_unrotated_pca <- scaled_hcw |>
  ordinate(~ principal(., nfactors = 2L, rotate = "none"))
biplot.psych(hcw_unrotated_pca)
```

Not only are the observations spread slightly more, but the data appears more evenly distributed compared to the previous biplots. 

Additionally, most variable vectors in the unrotated PCA biplot no longer load as clearly onto a single component. This makes it more challenging to determine the latent variables that explain the variance in our original variables. In this case, the study from which our data was obtained termed the two principal components _consideration_ and _structure_, respectively. That PCA as well as the FA we performed both allow us to easily make statements such as, "Question 4 on the questionnaire strongly assesses consideration, whilst question 12 strongly assesses structure." In a PCA without rotation, such interpretations are not as clear.

Ultimately, we see that FA is useful for obtaining a simply structure with interpretable latent variables. If we want a similar result that captures more variance and thus retains more information from the original data, we may opt for PCA followed by rotation. And if we want to capture more variance still, perhaps at the expense of interpretability, then unrotated PCA delivers that.

# Sources

1. Guastello, A. D., Brunson, J. C., Sambuco, N., Dale, L. P., Tracy, N. A., Allen, B. R., & Mathews, C. A. (2024). Predictors of professional burnout and fulfilment in a longitudinal analysis on nurses and healthcare workers in the COVID-19 pandemic. Journal of clinical nursing, 33(1), 288–303. https://doi.org/10.1111/jocn.16463
