---
title: "Extending ordr: FA"
output: 
  rmarkdown::html_vignette:
    toc: TRUE
vignette: >
  %\VignetteIndexEntry{Extending ordr: FA}
  %\VignetteEngine{knitr::rmarkdown}
  %\VignetteEncoding{UTF-8}
---

```{r, setup, include=FALSE}
knitr::opts_chunk$set(
  comment = '', fig.width = 6, fig.height = 6
)
```

# Introduction

This vignette demonstrates how to integrate methods for a new ordination class into **ordr.extra**. It follows the PCA vignette, which for reasons of underlying theory is a simpler case.
Since we'll refer to the prequel here occasionally, we recommend beginning there.

```{r, message = FALSE}
library(ordr)
library(psych)
```

## Factor Analysis

Factor analysis (FA) is a family of dimension-reducing methods of geometric data analysis. Unlike PCA, FA assumes the existence of latent variables (factors) behind a set of data, the number of which is less than the number of variables originally taken to be in the set. Like principal components, the factors in FA are linear combinations of the variables, and the variables are said to load onto the factors. However, factors are chosen to explain correlations between the variables, not to capture maximal variance; it is not assumed that factors are uncorrelated and orthogonal.

The core R distribution contains `stats::factanal()`, and **ordr** already recognizes its output. This vignette focuses on `psych::fa()`, with `factanal` methods as points of reference. Its integration into **ordr** will be much like that of `psych::principal()`, due to similarities in the linear algebra and in their interpretations, but several pitfalls must be avoided.

```{r, echo = TRUE, warning = FALSE, message = FALSE}
scaled_swiss <- scale(swiss)
fa_stats <- factanal(scaled_swiss, factors = 2, rotation = "none", scores = "regression")
(fa_tbl_ord <- as_tbl_ord(fa_stats))
```

## Factor Indeterminacy

FA assumes correlations between the factors, which leads to the issue of factor indeterminacy: Different optimization procedures may produce different sets of $k$ factors.
Moreover, the factors obtained are not ranked and may be inconsistent across different choices of $k$. This is in stark contrast to the fact that the first $l \leq k$ principal components are the same regardless of $k$.
Compare the results using $k=1$ and $k=2$:

```{r, echo = TRUE, warning = FALSE, message = FALSE}
fa(scaled_swiss, nfactors = 1)$loadings
fa(scaled_swiss, nfactors = 2)$loadings
```

## Matrix Decomposition

Factor analysis is based on the following model:
$$R_{p\times{p}} = L_{p\times{k}}(L_{p\times{k}})^\top + U_{p\times{p}}^2$$
Note that FA uses the correlation matrix $R$ rather than the covariance matrix, which can also be obtained by centering and scaling the data:
$$(\bar{X}_{n\times{p}})^\top\bar{X}_{n\times{p}} = L_{p\times{k}}(L_{p\times{k}})^\top + U_{p\times{p}}^2$$
As with PCA via EVD, we decompose this matrix into the product of a loadings matrix $L$ and its transpose. However, there is no diagonal matrix of inertia here, and we add to the decomposition a residuals matrix $U^2$.
We can check this on the decomposition returned by `psych::fa()`:

```{r}
# correlation matrix
cor_swiss <- cov(scaled_swiss)
# factor analysis
fa_psych <- 
  psych::fa(r = swiss, nfactors = 2L, rotate = "none", scores = "regression", fm = "ml")
# factor analysis loadings and residuals
L <- fa_psych$loadings
Usquared <- fa_psych$residual
# equivalence
cor_swiss / (L %*% t(L) + Usquared)
```

As mentioned above, the loadings matrix $L$ depends on the FA method; the `fm` argument selects among 12 methods provided by **psych**.
In this vignette, we'll hold the method fixed as we expand **ordr** around this new class, then verify that the expansion accommodates other methods correctly.

# Recovery Methods

We'll first assemble the recoverers one by one.
Then we'll write and test the `as_tbl_ord()` method for the payoff.

## Row and Column Elements

As with `psych::principal()` we take the factor loadings as the active column elements and the scores as supplementary row elements.
An additional complication in FA, from a PCA perspective, is that several methods are available to compute case scores, based on different choices of weights. The default, regression scores, uses the weight matrix $W_{p\times{k}} = R^{-1}_{p\times{p}}L_{p\times{k}}$.
The scores are obtained from the centered data matrix by multiplication on the right, $S_{n\times{k}} = \bar{X}_{n\times{p}}W_{p\times{k}}$.

<details>
<summary> Check how scores are computed </summary>

```{r}
# the matrix of 1s indicates equality
(w <- fa_psych$weights) / (solve(cor_swiss) %*% fa_psych$loadings)
# the matrix of 1s indicates equality
head(fa_psych$scores / (scaled_swiss %*% w))
```

</details>

The scores do not necessarily multiply with the loadings to recover the data, but we may obtain a matrix decomposition that justifies their inclusion as row elements.
The weight matrix $W_{p\times k}$ is not invertible, but the _pseudoinverse_ $W^+_{k\times p} = ((W_{p\times k})^\top W_{p\times k})^{-1}(W_{p\times k})^\top$ is a left inverse in the sense that $W^+W=I_{k\times k}$ when $W$ is full-rank and has the nice property that $WW^+W = W$.
We then have $S_{n\times k}W^+_{k\times p} = \bar{X}_{n\times p}W_{p\times k}W^+_{k\times p}$---that is, the scores multiply with the pseudoinverse weights to recover an idempotent transformation of the scaled data matrix.

<details>
<summary> Check how the pseudoinverse multiplies with the weights </summary>

```{r}
# define pseudoinverse W^+
w_inv <- solve(t(w) %*% w) %*% t(w)
# left and right products with W
w_inv %*% w
w %*% w_inv
# idempotency of right product
( w %*% w_inv %*% w ) / w
```

</details>

This decomposition situates the scores as row elements and also prompts us to include (the transpose of) the pseudoinverse weight matrix as a set of supplementary column elements.
Indeed, $(W^+_{k \times p})^\top$ and $L_{p\times k}$ share the same dimensions, and we can reason that both contain full inertia.

<details>
<summary> Reason that the same inertia is conferred on $(W^+)^\top$ as on $L$ </summary>

Let $d$ denote the inertia contained in $X$.
Then $(\bar{X}_{n\times p})^\top\bar{X}_{n\times p} = R_{p\times p}$ contains $d^2$. Recalling that $L_{p\times k}$ contains $d$, it follows that $W_{p\times{k}} = R^{-1}_{p\times{p}}L_{p\times{k}}$ contains $d^{-2}d = d^{-1}$.
Consequently, $W^+_{k\times p} = ((W_{p\times k})^\top W_{p\times k})^{-1}(W_{p\times k})^\top$ contains $(d^{-2})^{-1}d^{-1} = d^2d^{-1} = d$.
This is also consistent with the scores being in standard coordinates, as determined for `psych::principal()`.

</details>

We can now define the row and column recoverers, largely mimicking those of the `principal` class:

```{r}
# no active row elements
recover_rows.fa <- function(x) {
  matrix(
    nrow = 0, ncol = ncol(x[["loadings"]]),
    dimnames = list(NULL, colnames(x[["loadings"]]))
  )
}
# loadings as active column elements
recover_cols.fa <- function(x) {
  unclass(x[["loadings"]])
}
# scores as supplementary row elements
recover_supp_rows.fa <- function(x) {
  if (is.null(x[["scores"]])) {
    matrix(numeric(0), nrow = 0, ncol = 0)
  }
  else
    x[["scores"]]
}
# transpose pseudoinverse weight matrix as supplementary column elements
recover_supp_cols.fa <- function(x) {
  t( solve(t(x[["weights"]]) %*% x[["weights"]]) %*% t(x[["weights"]]) )
}
# illustrations
recover_cols(fa_psych)
recover_rows(fa_psych)
recover_supp_cols(fa_psych)
head(recover_supp_rows(fa_psych))
```

<details>
<summary> Check that elements agree with those from `factanal` </summary>

Let us benchmark these choices against our `factanal` object from earlier.
While `stats::factanal()` returns no supplementary columns, the remaining three sets of elements are approximately equal:

```{r}
recover_cols(fa_stats) / recover_cols(fa_psych)
recover_rows(fa_stats) / recover_rows(fa_psych)
head(recover_supp_rows(fa_stats) / recover_supp_rows(fa_psych))
```

</details>

### Conference of Inertia

As previously noted, the decomposition for FA does not have a distinct inertia matrix. This is a result of FA allowing correlations between factors. Whereas in PCA the uncorrelated structure of principal components mean that no two components can explain a portion of the same variance, the same is not true in FA. Hence, we cannot take eigenvalues as portions of explained variance in FA as we do in PCA.

To validate a conference of inertia for `psych::fa()`, we will use an alternative interpretation of inertia that can generalize from PCA to FA.

One meaning of the inertia in PCA is as the variances of the principal components. So, we can obtain the inertia by squaring each loadings entry, then taking the column sums, and lastly multiplying by $n−1$. Equivalence between this and the previous eigenvalue method of extracting inertia can be demonstrated using the code below.

```{r, eval = FALSE}
recover_inertia.principal(pca_psych)
colSums(pca_psych$loadings^2) * (nrow(scaled_iris) - 1)
```

Onto the proof for `psych::fa()`. We claim the function has an intertia conference of (0, 1).

First, we show that the scores multiplied by the transposed loadings gives an approximation of the data matrix.

```{r}
# create "fa" object, now with six factors
fa_psych6 <- psych::fa(swiss, nfactors = 6, rotate = "none")

stdized_approx <- fa_psych6$scores %*% t(as.matrix(fa_psych6$loadings))
means <- colMeans(swiss)
sds <- apply(swiss, 2, sd)
swiss_approx <- sweep(stdized_approx, 2, sds, "*") # de-scale
swiss_approx <- sweep(swiss_approx, 2, means, "+") # de-center

head(swiss_approx / swiss) # values close to 1 indicate near equality
```

Therefore, the data matrix is approximately the product of the scores and tranposed loadings matrices (after we undo centering and scaling). Then to prove that the scores contain zero inertia, it suffices to show that the loadings contain full inertia.

By default, the `cor` argument in `psych::fa()` is set to "cor", meaning the function performs FA on the correlation matrix. Thus, the total variance is 1 for each variable, i.e. 6 in this case.

Additionally, the total variance is equal to the sum of shared variance and uniquenesses. Observe below that the inertia from the loadings (calculated via column sums of squared entries) can be added to the uniquenesses to obtain the number of variables, as desired.

```{r}
common_var <- (colSums(fa_psych$loadings^2)) # inertia from loadings
uniq <- fa_psych$uniquenesses # uniquenesses

sum(common_var) + sum(uniq) # shared variance + uniquenesses
```

This implies that the loadings carry full inertia (minus uniquenesses). As previously shown, it follows that the scores carry zero inertia. So, we conclude that `psych::fa()` indeed has an inertia conference of (0, 1).

```{r}
recover_conference.fa <- function(x) {
  c(0, 1)
}
```

Seeing that the column sums of squares agree with the extracted uniquenesses in summing to total variance, we can also define our inertia recoverer as follows.

```{r}
recover_inertia.fa <- function(x) {
  colSums(x[["loadings"]] ^ 2)
}

recover_inertia.fa(fa_psych)
```

Then let us verify that these values returned by `recover_inertia.fa()` approximately equal those recovered from our `factanal` object.

```{r}
recover_inertia(fa_stats)
```

### Augmentation

```{r}
library(tibble)
```

Finally, we piece these recovered elements back into an augmented matrix for the user. As with PCA, we want to ascribe each element to a coordinate name, column/row attribute, and active/supplementary attribute. Additionally, we wish to present the user with the uniqueness of each variable as this element is directly accessible from the output of `psych::fa()` and indicates which variables account for the most variance. We write the following functions.

```{r}
recover_coord.fa <- function(x) {
  colnames(x[["loadings"]])
}

recover_aug_coord.fa <- function(x) {
  tibble(
    name = ordr:::factor_coord(recover_coord(x))
  )
}

recover_aug_rows.fa <- function(x) {
  res <- tibble(.rows = 0L)
  
  # scores as supplementary points
  if (!is.null(x[["scores"]])) {
    name <- rownames(x[["scores"]])
    res_sup <- if (is.null(name)) {
      tibble(.rows = nrow(x[["scores"]]))
    } else {
      tibble(name = name)
    }
  } else {
    res_sup <- tibble(.rows = 0L)
  }
  
  # supplement flag
  res$.element <- "active"
  res_sup$.element <- "score"
  as_tibble(dplyr::bind_rows(res, res_sup))
}

recover_aug_cols.fa <- function(x) {
  name <- rownames(x[["loadings"]])
  res <- if (is.null(name)) {
    tibble(.rows = nrow(x[["loadings"]]))
  } else {
    tibble(name = name)
  }
  res$uniqueness <- x$uniquenesses
  res$communality <- x$communality
  res$complexity <- x$complexity
  
  # supplement flag
  res$.element <- "active"
  # reorder columns
  res <- res[c(".element", setdiff(names(res), ".element"))]
  
  # transposed pseudoinverse of weights as supplementary points
  name <- rownames(x[["weights"]])
  res_sup <- if (is.null(name)) {
    tibble(.rows = nrow(x[["weights"]]))
  } else {
    tibble(name = name)
  }
  
  # supplement flag
  res_sup$.element <- "inv_weight"
  # reorder columns
  res_sup <- res_sup[c(".element", setdiff(names(res_sup), ".element"))]
  as_tibble(dplyr::bind_rows(res, res_sup))
}
```
We will verify that these augmentations have the expected number of rows.

```{r}
recover_aug_coord.fa(fa_psych) |> 
  nrow() # should yield two rows for the two factors

recover_aug_rows.fa(fa_psych) |>
  nrow() # should yield 47 rows for the 47 rows in the data frame/scores

recover_aug_cols.fa(fa_psych) |>
  nrow() # should yield twelve rows for the six rows of weights and six rows of loadings
```

So, all augmentations indeed give the correct number of rows. And thus, we can generate the below `tbl_ord` object with ease after defining `as_tbl_ord.fa()` as we did `as_tbl_ord.principal()`.

```{r}
as_tbl_ord.fa <- getFromNamespace("as_tbl_ord_default", "ordr")

as_tbl_ord(fa_psych) |>
  augment_ord()
```

### Examples

As with our recoverers for `psych::principal()`, we need to organize those for `psych::fa()` in a "methods" script. This time, we follow the format of `methods-stats-factanal.r`.

For our examples, we replicate those of `ex-methods-factanal-swiss.r` as follows:

```{r}
# data frame of Swiss fertility and socioeconomic indicators
class(swiss)
head(swiss)
# perform factor analysis
swiss_fa <- psych::fa(r = swiss, nfactors = 2L, rotate = "varimax", scores = "regression", 
               fm = "ml")

# wrap as a 'tbl_ord' object
(swiss_fa <- as_tbl_ord(swiss_fa))

# recover loadings
get_cols(swiss_fa, elements = "active")
# recover scores
head(get_rows(swiss_fa, elements = "score"))

# augment column loadings with uniquenesses
(swiss_fa <- augment_ord(swiss_fa))
```

### Unit Tests

Similarly, we apply the tests from `test-stats-factanal.r` to our `psych::fa()` recoverers for the unit test script. Note that, in **ordr.extra**, unit tests for GDA methods of the same package are included in a single script. In this case, we write the following unit tests for `psych::fa()` after the previous unit tests for `psych::principal()` in one file, `test-psych.R`.

```{r}
library(testthat)

fit_fa <- psych::fa(r = swiss, nfactors = 2L, rotate = "varimax", scores = "regression", 
                    fm = "ml")

test_that("'fa' accessors have consistent dimensions", {
  expect_equal(ncol(get_rows(fit_fa)), ncol(get_cols(fit_fa)))
  expect_equal(ncol(get_rows(fit_fa)),
               length(recover_inertia(fit_fa)))
})

test_that("'fa' has specified distribution of inertia", {
  expect_type(recover_conference(fit_fa), "double")
  expect_vector(recover_conference(fit_fa), size = 2L)
})

test_that("'fa' augmentations are consistent with '.element' column", {
  expect_equal(".element" %in% names(recover_aug_rows(fit_fa)),
               ".element" %in% names(recover_aug_cols(fit_fa)))
})

test_that("`as_tbl_ord()` coerces 'fa' objects", {
  expect_true(valid_tbl_ord(as_tbl_ord(fit_fa)))
})
```

### Submission

As in our integration of `psych::principal()` , we should conclude by running `devtools::check()`, then submit a pull request following the guidelines of `CONTRIBUTING.md`.
